### 2017-06-08 mediation.Rmd
start of mediation example. Perfect mediation probably won't work.Too much multicolinearity.

---
title: "Mediation"
author: "Ryan Gan"
date: "6/7/2017"
output: html_document
---

```{r load libraries, warning=F, message=F}
library(broom) # for easy conversion of model results to dataframes
library(tidyverse) # data wrangling
```

Simulating a mediated dataframe where x -> m -> y, where x perfectly predicts m.

```{r sim mediated data}
# set df observations
n <- 500
# set seed
seed(303)

# making continous x
sim_data <- data_frame(x = rnorm(n, 6, 0.5)) %>% 
  # simulate where x perfectly predits m
  mutate(m = x, 
  # simulate a protect effect between x and y
         y = rbinom(n, 1, 1/(1+exp(-(3.5+-0.6*m)))))
```

First step is to establish the relationship between x and y

```{r xy association}
summary(sim_data)
# check mod estimates
# this is referered to the direct effect
mod <- tidy((glm(y~x, sim_data, family="binomial"(link="logit"))))
mod
# OR
round(exp(mod[2,2]),3)
# 95 CI
round(exp(c(mod[2,2]-1.96*mod[2,3], mod[2,2]+1.96*mod[2,3])),3)
```

Second step is establishing x -> m, and m -> y relationship.

```{r xm my association}
xm_mod <- tidy(glm(m~x, sim_data, family = "gaussian"(link="identity")))
# perfect relationship 
xm_mod

# my mod
my_mod <- tidy(glm(y~m, sim_data, family="binomial"(link="logit")))
my_mod

# OR
round(exp(my_mod[2,2]),3)
# 95 CI
round(exp(c(my_mod[2,2]-1.96*my_mod[2,3], my_mod[2,2]+1.96*my_mod[2,3])),3)
```

Now that each path of the DAG has been confirmed, we calculate the controled direct effect.

```{r cde}

cde_mod <- tidy(glm(y~ x + m, sim_data, family="binomial"(link="logit")))

cde_mod
```

### 2017-06-12 mediation.Rmd
I may have a merge conflict. I never pushed my computer version of the document...
Didn't do anything...
updated markdown script.

@@
 ---
-title: "Mediation"
+title: "Simple Mediation Example"
 author: "Ryan Gan"
@@
 library(tidyverse) # data wrangling
+library(MASS) # simulate correlations between two continous variables
 ```
 
-Simulating a mediated dataframe where x -> m -> y, where x perfectly predicts m.
+Simulating a mediated dataframe where x -> m -> y
 
@@
 # set seed
-seed(303)
+set.seed(303)
+
+# simulate strong correlation between x and m
+
+xm_data <- as_tibble(mvrnorm(n, mu = c(6, 4), Sigma = matrix(c(1,0.7,0.7,1), ncol = 2)))
+# assign col names
+colnames(xm_data) <- c("x", "m")
+# check to amek sure correlation is ~70%
+cor(xm_data)
 
 # making continous x
-sim_data <- data_frame(x = rnorm(n, 6, 0.5)) %>% 
-  # simulate where x perfectly predits m
-  mutate(m = x, 
-  # simulate a protect effect between x and y
-         y = rbinom(n, 1, 1/(1+exp(-(3.5+-0.6*m)))))
+sim_data <- xm_data %>% 
+  # simulate protective effect between x and y
+  mutate(y = rbinom(n, 1, 1/(1+exp(-(1+-0.6*m)))))
 ```
 
-First step is to establish the relationship between x and y
+Some summary stats.
+
+```{r outcome pr}
+summary(sim_data)
+
+prop.table(xtabs(~sim_data$y))
+```
+
+In our simulated data, our outcome makes up roughly 21% of our population. Thus OR will not approximate true risk.
+
+First step is to establish the Total Effect (TE) relationship between x and y.
 
 ```{r xy association}
-summary(sim_data)
 # check mod estimates
 # this is referered to the direct effect
-mod <- tidy((glm(y~x, sim_data, family="binomial"(link="logit"))))
-mod
+te_mod <- tidy((glm(y~x, sim_data, family="binomial"(link="logit"))))
+te_mod
+
+# going to start to fill in standard estimates and CIs
+std_est <- as.data.frame(matrix(NA, ncol=4, nrow=3))
+colnames(std_est) <- c("est", "rel_est", "lower_95", "upper_95")
+
 # OR
-round(exp(mod[2,2]),3)
+std_est[1,1] <- "total_effect"
+std_est[1,2] <- round(exp(te_mod[2,2]),3)
 # 95 CI
-round(exp(c(mod[2,2]-1.96*mod[2,3], mod[2,2]+1.96*mod[2,3])),3)
+std_est[1, 3:4] <- round(exp(c(te_mod[2,2]-1.96*te_mod[2,3],
+                               te_mod[2,2]+1.96*te_mod[2,3])),3)
 ```
 
-Second step is establishing x -> m, and m -> y relationship.
+Second step is establishing x -> m, and m -> y relationship. These two estimates are required to meet the assumption, and to estimate the indirect/mediated path.
 
-```{r xm my association}
+```{r xm and my association}
+# linear x -> m mod beta
 xm_mod <- tidy(glm(m~x, sim_data, family = "gaussian"(link="identity")))
-# perfect relationship 
 xm_mod
 
-# my mod
+# logit m -> y mod beta
 my_mod <- tidy(glm(y~m, sim_data, family="binomial"(link="logit")))
@@
 
-Now that each path of the DAG has been confirmed, we calculate the controled direct effect.
+Now that each path of the DAG has been confirmed, we calculate the controled direct effect (CDE).
 
 ```{r cde}
+cde_mod <- tidy(glm(y~ x + m, sim_data, family="binomial"(link="logit")))  
+cde_mod
 
-cde_mod <- tidy(glm(y~ x + m, sim_data, family="binomial"(link="logit")))
+# OR
+std_est[2,1] <- "cde"
+std_est[2,2] <- round(exp(cde_mod[2,2]),3)
+# 95 CI
+std_est[2,3:4] <- round(exp(c(cde_mod[2,2]-1.96*cde_mod[2,3],
+                              cde_mod[2,2]+1.96*cde_mod[2,3])),3)
+```
 
+Notice the association betwene x and y essentially is null when m is in the model.
+Some old epi text books say this is evidence of mediation, assuming m is an intermediate variable.
+This might be a decent hint, but we can actually quantify the indirect effect, sometimes refered to Natural Indirect Effect (NIE) when there is no interaction between the mediator and predictor.
+
+When all our variables are continous TE - CDE = xm*teta2 = NIE
+
+Where theta2 is m on y effect in the adjusted CDE model.
+
+If all our variables are not continoous, then TE - CDE only approximates xm*theta2 = NIE. I'd advise use of the product method for most cases.
+
+```{r nie estimates}
+# te - cde
+nie1 <- round(te_mod[2,2] - cde_mod[2,2], 3) 
+# OR
+exp(nie1)
 cde_mod
+# xm * my nie
+nie2 <- round(xm_mod[2,2]*cde_mod[3,2],3)
+# OR
+exp(nie2)
 ```
+
+General interpretation for the NIE:
+
+A 1 unit increase in x as it passes through mediator m reduces the odds of y by 40%.
+
+Next step is to estimate 95% confidence intervals.
+We have two choices, the product method/Sobel?
+test and bootstrapping.
+In this simulated example, they should be equivalent.
+For small sample sizes like with the 9health data, it's better to bootstrap as the Sobel test is likely unstable.
+
+```{r nie se sobel}
+# we will use estimates from these models to caculate the joint se
+beta1 <- xm_mod[2,2]
+theta2 <- cde_mod[3,2]
+var_beta1 <- xm_mod[2,3]^2
+var_theta2 <- cde_mod[3,3]^2
+
+# calculate standard error of the nie
+nie_se <- sqrt((theta2^2*var_beta1)+(beta1^2*var_theta2))
+
+# estimate OR and 95% CI
+std_est[3,1] <- "nie"
+std_est[3,2:4] <- round(exp(c(nie2, nie2-1.96*nie_se, 
+                                         nie2+1.96*nie_se)),3)
+
+std_est
+```
+
+With small sample sizes, better to bootstrap bounds. I tend to use the percentile method since it's just easier to use with the broom package.
+
+```{r mediation function}
+# define function, mediation (simple function that doesn't have covariate)
+mediation <- function(data, y, x, m, y_family, m_family){
+
+ total_effect <- glm(y~x, data, family=y_family)$coefficients[2]
+ cde <- glm(y~x+m, data, family=y_family)$coefficients[2]
+ nie <- (glm(m~x, data, family=m_family)$coefficients[2]*
+         glm(y~x+m, data, family=y_family)$coefficients[3])
+
+ estimates <- cbind(total_effect, cde, nie)
+ return(estimates)
+
+}
+```
+
+Testing mediation function.
+
+```{r testing mediation funciton}
+# testing mediation function  
+test <- mediation(sim_data,y="y", x="x", m="m", y_family="binomial",
+                 m_family="gaussian")
+
+test
+```
+
+Bootstrapped estimate terms. Median, 2.5 and 97.5 percentile should be good approximations of mean and 95% CI estimate terms.
+
+```{r mediation boot}
+# booting 1000 times 
+boot_mediation <- sim_data %>% bootstrap(1000) %>% 
+  do(tidy(mediation(data=., y="y", x="x", m="m", y_family="binomial",
+                 m_family="gaussian")))
+
+# estimates (empty matrix)
+est <- matrix(NA, ncol=4, nrow=3) 
+colnames(est) <- c("est", "median", "lower_95", "upper_95")
+
+# fill total effect values
+est[1,1] <- "total_effect"
+est[1,2] <- round(exp(quantile(boot_mediation$total_effect, 0.5)),3)
+est[1,3] <- round(exp(quantile(boot_mediation$total_effect, 0.025)),3)
+est[1,4] <- round(exp(quantile(boot_mediation$total_effect, 0.975)),3)
+
+# fill cde values
+est[2,1] <- "cde"
+est[2,2] <- round(exp(quantile(boot_mediation$cde, 0.5)),3)
+est[2,3] <- round(exp(quantile(boot_mediation$cde, 0.025)),3)
+est[2,4] <- round(exp(quantile(boot_mediation$cde, 0.975)),3)
+
+# fill nie values
+est[3,1] <- "nie"
+est[3,2] <- round(exp(quantile(boot_mediation$nie, 0.5)),3)
+est[3,3] <- round(exp(quantile(boot_mediation$nie, 0.025)),3)
+est[3,4] <- round(exp(quantile(boot_mediation$nie, 0.975)),3)
+
+booted_estimates <- as_data_frame(est)
+
+# booted estimates
+knitr::kable(booted_estimates, caption = "Booted Estimated Effects")
+
+# standard estimates
+knitr::kable(std_est, caption = "Standard Estimated Effects")
+```

### 2017-06-12 mediation.Rmd
simple mediation function now handles covariates.

@@
 # simulate strong correlation between x and m 
-
-xm_data <- as_tibble(mvrnorm(n, mu = c(6, 4), Sigma = matrix(c(1,0.7,0.7,1), ncol = 2)))
+# also going to add a confounder correlated with both x and m
+cov_mat <- matrix(c(1,0.7,0.4,0.7,1,0.3,0.4,0.3,1), ncol=3)
+xm_data <- as_tibble(mvrnorm(n, mu = c(6, 4, 10), Sigma = cov_mat))
 # assign col names
-colnames(xm_data) <- c("x", "m")
+colnames(xm_data) <- c("x", "m", "c")
 # check to amek sure correlation is ~70%
@@
   # simulate protective effect between x and y
-  mutate(y = rbinom(n, 1, 1/(1+exp(-(1+-0.6*m)))))
+  mutate(y = rbinom(n, 1, 1/(1+exp(-(1+-0.6*m + 0.01*c)))))
 ```
@@
 # define function, mediation (simple function that doesn't have covariate)
-mediation <- function(data, y, x, m, y_family, m_family){
+mediation <- function(data, y, x, m, c, y_family, m_family){
 
- total_effect <- glm(y~x, data, family=y_family)$coefficients[2]
- cde <- glm(y~x+m, data, family=y_family)$coefficients[2]
- nie <- (glm(m~x, data, family=m_family)$coefficients[2]*
-         glm(y~x+m, data, family=y_family)$coefficients[3])
+  # if c covariates not specified
+  if(missing(c)){
+  # define total effect model
+  te_mod <- glm(y~x, data, family=y_family)
+  # define cde mod
+  cde_mod <- glm(y~x+m, data, family=y_family)
+  # define x m mod
+  m_x_mod <- glm(m~x, data, family=m_family)
+
+  # now if c is specified, adjust for covariates
+  } else {  
+  te_mod <- glm(as.formula(paste(y, paste(x,c, sep="+"), sep="~")), data,
+                family=y_family)
+
+  cde_mod <- glm(as.formula(paste(y, paste(x,m,c, sep="+"), sep="~")), data,
+                family=y_family)
+
+  m_x_mod <- glm(as.formula(paste(m, paste(x,c, sep="+"), sep="~")), data,
+                family=m_family)
+  } # end else
+
+  # fill in estimates from functions
+  total_effect <- te_mod$coefficients[2]
+  cde <- cde_mod$coefficients[2]
+  nie <- m_x_mod$coefficients[2]*cde_mod$coefficients[3]
 
@@
 ```{r testing mediation funciton}
-# testing mediation function  
+# testing mediation function without covariate
 test <- mediation(sim_data,y="y", x="x", m="m", y_family="binomial",
@@
 test
+
+# test with covariate
+test2 <- mediation(sim_data,y="y", x="x", m="m", c="c", y_family="binomial",
+                 m_family="gaussian")
+
+test2
 ```
@@
 boot_mediation <- sim_data %>% bootstrap(1000) %>% 
-  do(tidy(mediation(data=., y="y", x="x", m="m", y_family="binomial",
+  do(tidy(mediation(data=., y="y", x="x", m="m", c="c", y_family="poisson",
                  m_family="gaussian")))

### 2017-06-12 mediation.Rmd
added note that I haven't updated the rest of the text in the document.
updated step by step to match function estimates.

@@
 # this is referered to the direct effect
-te_mod <- tidy((glm(y~x, sim_data, family="binomial"(link="logit"))))
+te_mod <- tidy((glm(y~x+c, sim_data, family="poisson"(link = "log"))))
 te_mod
@@
 # linear x -> m mod beta
-xm_mod <- tidy(glm(m~x, sim_data, family = "gaussian"(link="identity")))
+xm_mod <- tidy(glm(m~x+c, sim_data, family = "gaussian"(link="identity")))
 xm_mod
@@
 # logit m -> y mod beta
-my_mod <- tidy(glm(y~m, sim_data, family="binomial"(link="logit")))
+my_mod <- tidy(glm(y~m+c, sim_data, family="poisson"(link="log")))
 my_mod
@@
 ```{r cde}
-cde_mod <- tidy(glm(y~ x + m, sim_data, family="binomial"(link="logit")))  
+cde_mod <- tidy(glm(y~ x + m + c, sim_data, family="poisson"(link="log")))  
 cde_mod
@@
 ```{r mediation boot}
+set.seed(321)
 # booting 1000 times 
@@
 ```
+
+### TO DO
+
+I haven't modified the rest of the document after updating the funciton to handle confounding.

### 2017-06-13 mediation.Rmd
updated function to include relationship between x and m.

@@
   nie <- m_x_mod$coefficients[2]*cde_mod$coefficients[3]
+  x_affect_on_m <- m_x_mod$coefficients[2]
 
-  estimates <- cbind(total_effect, cde, nie)
+  estimates <- cbind(total_effect, cde, x_affect_on_m, nie)
   return(estimates)
@@
 # estimates (empty matrix)
-est <- matrix(NA, ncol=4, nrow=3) 
+est <- matrix(NA, ncol=4, nrow=4) 
 colnames(est) <- c("est", "median", "lower_95", "upper_95")
@@
 
+# fill m x relationship
+est[4,1] <- "x_m_association"
+est[4,2] <- round(exp(quantile(boot_mediation$x_affect_on_m, 0.5)),3)
+est[4,3] <- round(exp(quantile(boot_mediation$x_affect_on_m, 0.025)),3)
+est[4,4] <- round(exp(quantile(boot_mediation$x_affect_on_m, 0.975)),3)
+
 booted_estimates <- as_data_frame(est)

### 2017-06-13 mediation.Rmd
added some additional estimates to help interpretation of mediation.

@@
   nie <- m_x_mod$coefficients[2]*cde_mod$coefficients[3]
-  x_affect_on_m <- m_x_mod$coefficients[2]
+  y_on_m <- cde_mod$coefficients[3]
+  x_on_m <- m_x_mod$coefficients[2]
 
-  estimates <- cbind(total_effect, cde, x_affect_on_m, nie)
+  estimates <- cbind(total_effect, cde, nie, x_on_m, y_on_m)
   return(estimates)

### 2017-06-20 categorical_msm.Rmd
adding in categorical msm code

---
title: "Categorical MSM"
author: "Ryan_Gan"
date: "June 7, 2017"
output: html_document
---

```{r libraries, message=F}
library(tidyverse)
library(broom)
library(contrast)
```

## Simulated dataset

Simulating a simple continous distribution and a linear relationship with binary outcome which I'll then break in to an ordinal catogory by tertiles. No confounding yet.

```{r sim data}
n <- 500
set.seed(333)

# create normal distribution of x with mean of 16 and sd of 4
sim_data <- data_frame(x = rnorm(n, 16, 4)) %>% 
  # simulate y data with linear logit relationship to x
  mutate(y = rbinom(n, 1, 1/(1+exp(-(-1+0.10*x)))),
         # split in to three categories
         x_cat = ifelse(x < quantile(x,1/3), 1,
                 ifelse(x > quantile(x,2/3), 3, 2)),
         # dummy variable
         x1 = ifelse(x_cat == 1, 1, 0),
         x2 = ifelse(x_cat == 2, 1, 0),
         x3 = ifelse(x_cat == 3, 1, 0),
         x2_3 = ifelse(x_cat == 2 | x_cat == 3, 1, 0))
```

```{r standard estimates}
# check density plots of x category given x
ggplot(sim_data, aes(x=x, fill=as.factor(x_cat))) +
  geom_density(alpha = 0.5)

# estimates good enough for example
mod1 <- glm(y~x, sim_data, family="binomial"(link="logit"))
mod1$coefficients

# plot predicted probability of the linear funciton
pr_y_x <- predict(mod1, sim_data, type="response")

# plot of predicted probability of continous variable
ggplot(sim_data, aes(x=x, y=pr_y_x)) +
  geom_point()

```

Standard analysis asking what the risk difference is strata to strata

```{r risk difference}
# 2x3 table
y_x <- xtabs(~x_cat+y, sim_data)
pr_yx <- prop.table(y_x, 1)
pr_yx
# difference in risk from strata 2 to 1
pr_yx[2,2] - pr_yx[1,2]
# difference in risk from strata 3 to 1
pr_yx[3,2] - pr_yx[1,2]
# difference in risk from strata 3 to 2
pr_yx[3,2] - pr_yx[2,2]

# estimate risk difference using binomial, link=identity for appropriate SE
risk_diff_mod <- glm(y~x2+x3, sim_data, family="binomial"(link="identity"))
summary(risk_diff_mod)

# comparing difference in x3 to x2
risk_diff_mod2 <- glm(y~x3+x1, sim_data, family="binomial"(link="identity"))
summary(risk_diff_mod2)
# difference between groups 3 and 2 as well
# now what if we combine groups x 3 and 2 together and compared against 1

risk_diff_mod3 <- glm(y~x2_3, sim_data,family="binomial"(link="identity"))
summary(risk_diff_mod3)

```

### 2017-06-20 categorical_msm.Rmd
updated

@@
 library(tidyverse) 
-library(broom)
+library(broom) # for tidy model output and bootstrapping
+library(MSModelR) # my package for simple msm functions
 library(contrast)
@@
          # split in to three categories
-         x_cat = ifelse(x < quantile(x,1/3), 1,
-                 ifelse(x > quantile(x,2/3), 3, 2)),
+         x_cat = ifelse(x < quantile(x,1/3), 0,
+                 ifelse(x > quantile(x,2/3), 2, 1)),
          # dummy variable
+         x0 = ifelse(x_cat == 0, 1, 0),
          x1 = ifelse(x_cat == 1, 1, 0),
          x2 = ifelse(x_cat == 2, 1, 0),
-         x3 = ifelse(x_cat == 3, 1, 0),
-         x2_3 = ifelse(x_cat == 2 | x_cat == 3, 1, 0))
+         x1_2 = ifelse(x_cat == 1 | x_cat == 2, 1, 0))
 ```
+
+Truth.
 
@@
 # estimates good enough for example
-mod1 <- glm(y~x, sim_data, family="binomial"(link="logit"))
-mod1$coefficients
+true_mod <- glm(y~x, sim_data, family="binomial"(link="logit"))
+tidy(true_mod)
 
 # plot predicted probability of the linear funciton
-pr_y_x <- predict(mod1, sim_data, type="response")
+pr_y_x <- predict(true_mod, sim_data, type="response")
 
@@
 
-Standard analysis asking what the risk difference is strata to strata
+Standard analysis asking what the risk ratio is strata to strata
 
-```{r risk difference}
+```{r risk ratio}
 # 2x3 table
@@
 
-# estimate risk difference using binomial, link=identity for appropriate SE
-risk_diff_mod <- glm(y~x2+x3, sim_data, family="binomial"(link="identity"))
-summary(risk_diff_mod)
+# estimate risk ratio
+risk_ratio_mod <- glm(y~x1+x2, sim_data, family="binomial"(link="logit"))
+tidy(risk_ratio_mod)
 
-# comparing difference in x3 to x2
-risk_diff_mod2 <- glm(y~x3+x1, sim_data, family="binomial"(link="identity"))
-summary(risk_diff_mod2)
-# difference between groups 3 and 2 as well
+# comparing difference in x2 to x1
+risk_ratio_mod2 <- glm(y~x2+x0, sim_data, family="binomial"(link="logit"))
+tidy(risk_ratio_mod2)
+
+# treating the categorical variables as continous
+risk_ratio_mod3 <- glm(y~x_cat, sim_data, family="binomial"(link="logit"))
+tidy(risk_ratio_mod3)
+
 # now what if we combine groups x 3 and 2 together and compared against 1
+risk_ratio_mod4 <- glm(y~x1_2, sim_data,family="binomial"(link="logit"))
+tidy(risk_ratio_mod4)
+```
 
-risk_diff_mod3 <- glm(y~x2_3, sim_data,family="binomial"(link="identity"))
-summary(risk_diff_mod3)
+Loss of information by combining the mid and high strata.
+But would this truly reflect a population effect where we may ask the question what if we reduced to background levels represented by strata 0?
+Something to think about.
 
-```
+Trying out MSM estimates.

### 2017-07-12 msm_introduction.Rmd
added msm presentation to repo.

---
title: "Introduction to Causal Inference Models"
author: "Ryan Gan"
date: "7/10/2017"
output: slidy_presentation
---

## Setup

I will use the tidyverse and broom package for this presentation.

```{r setup, message=F, warning=F}
library(tidyverse)
library(broom)
```

## "Causal Inference"

If you've been to Society for Epidemiologic Research or any epi methods conference, you've probably heard something about "causal inference".

You've also probably heard a lot of gibberish about g-estimation, g-computation, g-formula, g-something, inverse-probability-of-treatment weighting, marginal structural models, doubly-robust, targeted maximum likelihood, etc.

And like me, you probably have no idea what any of those are, what they are used for, or even how to use them. But they are always referred to as being used in "causal inference", so these fancy-pants models must be special and the only way to find causal associations.

This presentation provides a general overview of things I've learned about these methods. I'm still confused about a lot of this topic area (like how is the g-formula different from g-estimation?), but hopefully this will server as a primer.

## Causal Inference Models

These models help us answer the marginal effects of a binary exposure and help us deal with assessing the marginal effect of a binary exposure that varies over time, and is also confounded over time (original motivation of Robins). These methods can be applied to continous or binary outcomes, but only binary exposures.

- Two general categories that have been around for a decade or so

1. Structural nested models 
- Commonly referred to as g-estimation, g-computation, g-formula, g-something
- The g stands for "general"

2. Marginal structural models 
- Sometimes called inverse-probability-of-treatment weighting or IPTW

- Newer approach in recent years are the doubly-robust method which we will touch on later

## A Mostly Accurate History of Causal Models

A long time ago (1920s), Jerzy Neyman proposed a class of statistical models that would latter be referred to as counter factual/potential outcomes models for randomized trials. Donald Rubin extended this framework to observational data and coined the term potential outcomes in maybe the 1970s?. Judea Pearl and Jamie Robins also did some stuff in the 80's. Jamie Robins introduced the g-formula and IPTW which were designed to estimate the marginal effect of a binary treatment that varied over time and was confounded by a time-varying confounder. Jamie Robins and a lot of his students have made some tweaks to the original framework which pretty much brings us to today.

Anyways, a major process/step to most of these early models is the concept of the counter factual (which we'll cover in depth later in this presentation).

The general concept of the counter factual is it takes our "what if?" questions from implicit to explicit in our data.

## Counterfactuals: Making Explicit Causal Claims

- The randomized control trial (RCT): Our gold standard for assessing causal relationships. The basic study design has the investigators randomly assign a binary treatment to a population and follow the population over time to identify incident outcomes. Any potential confounding variables are distributed evenly among our treatment group, which creates d-separation/independence/no association between our confounding variables and our treatment. With this study design, we can estimate causal and unbiased effects between treatment and outcomes.

- What is implicit in this design is that the 'no' treatment group represents what would have happened to the 'yes' treatment group had they not been given the treatment.

- Counter factual estimation makes this implicit assumption explicit.

## Counterfactual Estimation is a Problem of Missing Data

- The problem with any epidemiologic study design is that a subject's exposure (denoted as A) and outcome (denoted as Y) can be observed only once.

```{r simple simulation}
# set sample size and seed 
n <- 5000; set.seed(777)
# simulate exposure/treatment data with equal probability
sim_data <- as_tibble((rbinom(n, 1, prob = 0.5))) %>% 
  rename(A = value) %>% 
  # simulate protective effect between A and Y
  mutate(Y = rbinom(n, 1, 1/(1+exp(-(-2.5 + -0.6*A)))))
# print first 6 observations
head(sim_data)
```

- For example, our first subject has been treated and has not developed the outcome.

## Counterfactual Estimate using G-Computation

- G-computation is a way in which we can estimate the counter factual state where everyone is treated/exposed (denoted by a = 1) and no one is treated/exposed (denoted by a = 0)

- G-computation is implemented in a 3-step process:

*1. Estimation of the conditional mean outcome E(Y|A)*

*2. Obtaining predicted outcomes under a = 1 and a = 0*

*3. Estimate the statistical parameter (difference/ratio) by substituting the predicted mean outcomes under a = 1 and a = 0*

## Estimation of the Conditional Mean Outcome E(Y|A)

This can be estimated with your standard generalized linear model. We'll estimate the risk ratio, but risk difference can be estimated as well.

*2x2 Table*
```{r 2x2}
xtabs(~A + Y, sim_data)
```

*Risk Ratio of E(Y|A)*

```{r ratio}
# standard ratio model
ratio_model <- glm(Y~A, sim_data, family = "binomial"(link="logit"))
# ratio estimate
exp(ratio_model$coefficients[2])
```
The treated group is 40% less likely to develop the outcome compared to the untreated group.

## Updating the Predicted Outcomes Under a = 1

First we duplicate our original dataframe and then set everyone's treatment to 1.

```{r create a1 exposure dataframe}
# set everyone to exposed from our original dataframe
sim_data_a1 <- sim_data %>% mutate(A = 1)
# view first 6 records
head(sim_data_a1)
```

Now we apply our saved model coefficients (ratio_model) and updated the predicted values using the dataframe where everyone receives treatment E(Y|a=1).

```{r pr a1}
# updated predicted probability of outcome 
pr_a1 <- predict(ratio_model, newdata = sim_data_a1, type = "response")
# average probability where everyone is exposed
mean(pr_a1)
```

## Updating the Predicted Outcomes Under a = 0

We repeat the same step under the scenario where no one is treated.

```{r create a1 exposure dataframe}
# set everyone to exposed from our original dataframe
sim_data_a0 <- sim_data %>% mutate(A = 0)
# view first 6 records
head(sim_data_a0)
```

Updating our predicted probability under no treatment scenario E(Y|a=0).

```{r pr a0}
# updated predicted probability of outcome 
pr_a0 <- predict(ratio_model, newdata = sim_data_a0, type = "response")
# average probability where no one is exposed
mean(pr_a0)
```

## Estimate the Counterfactual Parameter of Interest

Now that we have a mean value for Y where everyone is treated and no one is treated, we can estimate our risk ratio.

```{r g forumla ratio}
mean(pr_a1)/mean(pr_a0)
```

Let's compare to our estimate using logistic regression again.

```{r standard est result}
exp(ratio_model$coefficients[2])
```

We get pretty much the same result. That was a lot of steps to get the same answer.

### 2017-07-12 g_comp_example_snowden.Rmd
moved my scripts to this general repo about methods.

---
title: "Reproducing Results in Snowden G-Comp Paper"
author: "Ryan_Gan"
date: "April 18, 2017"
output: html_document
---

Using code from Snowden paper to try out MSM models and maybe test out SuperLearner and Target Maximum Likelihood packages.

```{r libraries, message=F}
library(tidyverse)
library(broom)
library(SuperLearner)
library(tmle)
```

Simulating data as defined in the paper. Paper goes in to various model fits and how MSM estimates. First step of a marginal structral model is to define the I'm just going to go to example 4 where the appropriate interaction between A and W2 is specified.

```{r simulate data}
# sample size
n <- 300
set.seed(285)
simdata <- data.frame(W1 = rbinom(n, 1, 0.4), W2 = rbinom(n, 1, 0.5))
# add simulated data of A
simdata <- transform(simdata, A = rbinom(n, 1, (0.5+0.2*W1-0.3*W2)))
# add simulated data of Y
simdata <- transform(simdata, Y = rnorm(n, (3-0.5*A+W1+0.3*A*W2), 0.4))

# create counterfactual datasets 
# A = 0
simdata_a0 <- transform(simdata, A = 0)
# A = 1
simdata_a1 <- transform(simdata, A = 1)
```

## G-computation (simple substitution)

Applying simple substitution method (g-computation).

*Step 1. Estimation of the conditional mean outcome Eo(Y|A,W,A:W2)*

```{r gcomp step 1}
# Step 1: Define Q formula
reg_mod <- glm(Y~A+W1+A:W2, data = simdata, family = gaussian)
summary(reg_mod)
```

*Step 2. Obtaining predicted outcomes under everyone exposed and no one exposured*

Next we need to update the predicted FEV1 values where everyone is exposed and no one is exposed.

```{r gcomp step 2}
# Step 2: Updated predicted value
# values of everyone exposed
pr_fev_a1 <- predict(reg_mod, newdata = simdata_a1)
# values of no one exposed
pr_fev_a0 <- predict(reg_mod, newdata = simdata_a0)
```

*Step 3. Estimate the statistical parameter by substituting the predicted mean outcomes under everyone exposed and no one exposed*

```{r gcomp step 3}
mean(pr_fev_a1)
mean(pr_fev_a0)
# MSM value
fev_est <- mean(pr_fev_a1) - mean(pr_fev_a0)
fev_est
```

In order to get estimates of uncertainty around our MSM estimate ( 95% confidence invervals), we need to bootstrap.

First I create a general function that calculates g-comp with a couple user-defined inputs.

```{r gcomp function and bootstrap}
# might be able to cut out y element of formula and add it to q_mod

# create a g-comp function
g_comp_func <- function(data, y, a, q_mod, mod_family, estimate){
  # create a = 1 and a = 0 dataframes ----
  a1_df <- a0_df <- data
  a1_df[, a] <- 1
  a0_df[, a] <- 0
  # run q-formula ----
  # may consider moving model specification outside of function
    model <- glm(as.formula(paste(y, q_mod, sep = "~")), data = data, family = mod_family)
     # calculate mean expected in a1 and a0
      pr_a1 <- mean(predict(model, newdata = a1_df, type = "response")) 
      pr_a0 <- mean(predict(model, newdata = a0_df, type = "response"))

  # calculate desired estimates "diff", "ratio", "both" ----
    # calculate difference a1 to a0
    if(estimate == "diff"){
      msm_diff <- pr_a1 - pr_a0
      msm_val <- msm_diff
    }
    # calculate ratio of a1/a0
    if(estimate == "ratio"){
      msm_ratio <- pr_a1/pr_a0 
      msm_val <- msm_ratio
    } 
    # output (i want to figure out how to assign it a name)
      return(msm_val)

    # calculate both ratio and difference
    # if(estimate == "both"){
    #   msm_diff <- pr_a1 - pr_a0
    #   msm_ratio <- pr_a1/pr_a0
    #   msm_val <- rbind(c("Risk Difference", msm_diff),
    #                    c("Risk Ratio", msm_ratio))
    # }

} # end function
```

Testing function to see if it produces the same results.

```{r gcomp func test}
gcomp_est <- g_comp_func(data = simdata, y = "Y", a = "A", q_mod = "A+W1+A:W2",
            mod_family = "gaussian", estimate = "diff")
gcomp_est
```

Function works. Now we can feed it through broom's bootstrap function and calculation 95%CIs via percentile method.

```{r gcomp boot}
# percentile method with broom
set.seed(321)
boot_msm <- simdata %>% bootstrap(1000) %>% 
  do(tidy(g_comp_func(data = ., y = "Y", a = "A", q_mod = "A+W1+A:W2",
            mod_family = "gaussian", estimate = "diff")))

# create empty matrix
estimates <- matrix(nrow = 1, ncol = 4, byrow = T)
colnames(estimates) <- c("msm_estimate", "boot_median", "lower_95", "upper_95")

# fill matrix 
estimates[,1] <- round(gcomp_est,3)
estimates[,2] <- round(quantile(boot_msm$x, 0.5),3)
estimates[,3] <- round(quantile(boot_msm$x, 0.025),3)
estimates[,4] <- round(quantile(boot_msm$x, 0.975),3)

# convert matrix to dataframe
est_df <- data.frame(estimates)

knitr::kable(est_df, caption = paste0("MSM estimates with bootstrapped 95%CI"))
```

## Inverse probability of treatment weight (IPTW)

This method is similar to g-computation, but differs slightly in that confounding can be thought of as a problem of bias sampling where certain exposure-covariate subgroups are over or under represted compared to an RCT. IPTW up-weights under-represted subjects from exposure-covariate strata. Sounds like Bayesian nonsense.

*Step 1. Estimate the propensity score Pr(A|W)*

I'm actually interested to see how this will work since Snowden et al put an interesting interaction in. Maybe this was the point of their paper? Need to read it again on why they used g-comp over IPTW.

```{r iptw step 1}
prop_mod <- glm(A ~ W1 + W2, family = "binomial", data = simdata)
summary(prop_mod)
```

*Step 2. Create weights*

Multi-step process of creating weights for the probability of being exposed. We also check for ETA violations here where, given the covariates W, probability of exposed or unexposed might be >.90.

IPTW requires the assumption of positivity or experimental treatment assignment where the observed treatment levels vary within confounder strata. Positivity violations occur when certain subgroups in a sample rarely or never receive some treatments of interest. The resulting sparsity in the data may increase bias with or without an increase in variance and can threaten valid inference.

```{r iptw step 2}
# vector of propensity exposed for each observation
prob_exposed <- predict(prop_mod, type = "response")
# vector of propensity not exposed
prob_unexposed <- 1 - prob_exposed 
# histogram to check for ETA violation
hist(prob_exposed)

# create weights
wt <- as.numeric(simdata$A == 1)/prob_exposed + 
  as.numeric(simdata$A == 0)/prob_unexposed

# checking to make sure we inverted probabilities correctly
head(data.frame(simdata$Y, simdata$A, 1/prob_exposed, 1/prob_unexposed, wt))
```

*Step 3. Estimate statistical parameter*

```{r iptw step 3}
iptw <- mean(as.numeric(simdata$A==1)*wt*simdata$Y) -
  mean(as.numeric(simdata$A==0)*wt*simdata$Y)

iptw
```

Creating IPTW funciton to run through bootstrap.

```{r iptw function}
# inverse probability of treatment weighting function
iptw_func <- function(data, y, a, w, estimate){
  # step 1: estimate propensity of treatment ----
    prop_mod <- glm(as.formula(paste(a, w, sep = "~")), data = data, 
                    family = "binomial")
  # step 2: estimate weights ----
    # vector of propensity exposed for each observation
    prob_exposed <- predict(prop_mod, type = "response")
    # vector of propensity not exposed
    prob_unexposed <- 1 - prob_exposed

    # note: 4/20/17 can't quite figure out how to include a plot and value
    # histogram to check for ETA violation 
    # if(eta_plot == "yes"){ # plot probability of exposed
    # eta_fig <- hist(prob_exposed)
    #   }

  # create weights
    wt <- as.numeric(data[,a]==1)/prob_exposed + 
          as.numeric(data[,a]==0)/prob_unexposed

  # calculate desired estimates "diff", "ratio", "both" ----
    # calculate mean estimate
    a1_est <- mean(as.numeric(data[,a]==1)*as.numeric(wt)*as.numeric(unlist(data[,y])))
    a0_est <-  mean(as.numeric(data[,a]==0)*as.numeric(wt)*as.numeric(unlist(data[,y])))
    # calculate difference a1 to a0
    if(estimate == "diff"){
      iptw_diff <- a1_est - a0_est
      iptw_val <- iptw_diff
    }
    # calculate ratio of a1/a0
    if(estimate == "ratio"){
      iptw_ratio <- a1_est/a0_est
      iptw_val <- iptw_ratio
    } 
    # output a list of figure and value
    #iptw_out <- list(eta_fig, iptw_val)
    return(iptw_val)
} # end function

```

Testing function. You can see that IPTW is slightly biased compared to the g-computation approach because it doesn't account for the interaction between A and W2 without the baseterm W2.

```{r testing iptw function}
iptw_est <- iptw_func(data=simdata, y="Y", a="A", w="W1+W2", estimate="diff")
iptw_est
```

```{r iptw boot}
# percentile method with broom
set.seed(321)
boot_iptw <- simdata %>% bootstrap(1000) %>% 
  do(tidy(iptw_func(data=., y="Y", a="A", w="W1+W2", estimate="diff")))

# create empty matrix
estimates <- matrix(nrow = 1, ncol = 4, byrow = T)
colnames(estimates) <- c("iptw_estimate", "boot_median", "lower_95", "upper_95")

# fill matrix 
estimates[,1] <- round(iptw_est,3)
estimates[,2] <- round(quantile(boot_iptw$x, 0.5),3)
estimates[,3] <- round(quantile(boot_iptw$x, 0.025),3)
estimates[,4] <- round(quantile(boot_iptw$x, 0.975),3)

# convert matrix to dataframe
est_df <- data.frame(estimates)

knitr::kable(est_df, caption = paste0("IPTW estimates with bootstrapped 95%CI"))
```

## Targeted Maximum Likelihood Substitution

This is a doubly robust estimation method that uses ensemble machine learning (via SuperLearner) to estimate both the Q function and treatment propensity. If even one of these

*Step 1: Estimate E(Y|A,W) using ensemble machine learning*

```{r tmle q mod, message=F}
# register superlearner library
sl_lib <- c("SL.glm", "SL.glm.interaction", "SL.gam", "SL.nnet", 
            "SL.randomForest", "SL.step")

# estimate FEV/Y given the covariates
sl_outcome_reg <- SuperLearner(Y=simdata$Y, X=subset(simdata, select = -Y),
                    SL.library = sl_lib, family = "gaussian")
# output of superlearner predictive algorithm
sl_outcome_reg
```

SuperLearner is an esemble method that weights each machine learning algorithm used. The lower the risk, the better. The Coef variable shows the weight each model is contributing to the overall ensemble. GLM interaction contributes the most weight. We'd kind of hope that since we have that weird interaction between A and W2 without base contribution of W2.

Now that we have esemble formula (which is hidden to my knowledge) to predict Y, we can updated our data.

*Step 2. Use SuperLearner to estimate y where everyone is exposed and no one is exposed*

```{r sl predict exp and unexp}
# use predict function to obtain initial estimates of the expected outcome given the
# observed exposure and covariates En(Y|A,W)
expected_y_obsexp <- predict(sl_outcome_reg, newdata=simdata)$pred

# like g-computation, we can also update the predicted probability where everyone is exposed
expected_y_a1 <- predict(sl_outcome_reg, newdata=simdata_a1)$pred
# and where no one is exposed
expected_y_a0 <- predict(sl_outcome_reg, newdata=simdata_a0)$pred

head(cbind(simdata, expected_y_obsexp, expected_y_a1, expected_y_a0))

```

*Step 3. Estimate MSM*

```{r sl msm}
tmle_msm <- mean(expected_y_a1) - mean(expected_y_a0)
tmle_msm
```

Not bad for a machine. You can always add more machine learning methods and add crossvalidation to the ensemble, but that would increase computation time by quite a bit.

## Doubly Robust Method

This method also uses a machine learning ensemble method (SuperLearner) to not only predict the optimal Q model, but also the propensity of treatment model. This method has the advantage of being doubly robust, where if either the Q model or propensity of treatment model is correct, we will produce unbiased results.

*Steps 1. Same as TMLE substitution method where we define the Q model*

I'm not going to go over this again since we just did this.

*Step 2. Estimate propensity score using ensemble method*

We follow a similar method to the IPTW framework, but now we can create a 'clever covariate'. For exposed subjects, the clever covariate is 1 over the predicted probability of being exposed, given baseline covariates. For unexposed subjects, the clever covariate is -1 over the predicted probablity of not being exposed, given the baseline covariates.

```{r sl pscore}
# estimate propensity score
sl_pscore <- SuperLearner(Y=simdata$A, X=subset(simdata, select = -c(A,Y)),
                  SL.library = sl_lib, family = "binomial")
sl_pscore

# predicted probability of beind exposed given baseline covariates
expected_a1 <- sl_pscore$SL.predict
# predicted probability of being unexposed given baseline covariates
expected_a0 <- 1 - expected_a1

# creation of clever covariate.
clever_cov <- as.numeric(simdata$A==1)/expected_a1 -
              as.numeric(simdata$A==0)/expected_a0

# check to make sure clever covariate works
clev_cov_a1 <- 1/expected_a1
clev_cov_a0 <- -1/expected_a0

head(data.frame(simdata$A, clever_cov, clev_cov_a1, clev_cov_a0))
```

*Step 3. Target intial estimator*

Here we update the intial Q model with the clever covariate as an offset and supress the intercept.

```{r tmle step 3}
# offset needs to be qlogis for binary 
# normal might not need anything since it's identity link?
fev_mod_update <- glm(simdata$Y ~ -1 + offset(expected_y_obsexp) + 
                        clever_cov, family = "gaussian")

summary(fev_mod_update)

# the clever covariate estimate on Y will become epsilon
epsilon <- fev_mod_update$coefficients
epsilon

summary(glm(simdata$Y ~ clever_cov, family = "gaussian"))

#pnorm(expected_y_obsexp)

# estimated Y|A=1, W and Y|A=0, W
tmle_y_a1 <- expected_y_a1+epsilon*clev_cov_a1
tmle_y_a0 <- expected_y_a0+epsilon*clev_cov_a0
```

*Step 4. Substitute the updated fits into the target parameter*

```{r doubly robust tmle step 4}
tmle_dr_est <- mean(tmle_y_a1) - mean(tmle_y_a0)
tmle_dr_est
```

### Comparison of all MSM estimate methods

```{r comparsion}
msm_fev <- cbind(gcomp_est, iptw_est, tmle_msm, tmle_dr_est)

msm_fev
```

### TMLE Package

Trying out the tmle package.

```{r tmle package}
set.seed(123)
tmle_package_result <- tmle(Y = simdata$Y, A = simdata$A, 
  W = subset(simdata, select = -c(A,Y)),
  Q.SL.library = c("SL.nnet", "SL.glm", "SL.randomForest", "SL.step",
        "SL.glm.interaction"), 
  g.SL.library = c("SL.nnet", "SL.glm", "SL.randomForest", "SL.step",
        "SL.glm.interaction"),  V = 10)

tmle_package_result
```

Wow. This was much faster and produced 95% CIs. TMLE has it's own formla for calculating SE as described in package description in J of Stats Software. It tells me the same thing as the MSM bootstrapped method.

### 2017-07-12 msm_introduction.Rmd
moved my scripts to this general repo about methods.

@@
 2. Marginal structural models 
-- Sometimes called inverse-probability-of-treatment weighting or IPTW
+- Sometimes called inverse-probability-of-treatment weighting (IPTW)
 
@@
 
-## A Mostly Accurate History of Causal Models
+## A Brief History of Causal Models
 
-A long time ago (1920s), Jerzy Neyman proposed a class of statistical models that would latter be referred to as counter factual/potential outcomes models for randomized trials.
+A long time ago (1920s), in a country far far away (Poland), Jerzy Neyman proposed a class of statistical models that would latter be referred to as counter factual/potential outcomes models for randomized trials.
 Donald Rubin extended this framework to observational data and coined the term potential outcomes in maybe the 1970s?.
@@
 
-Anyways, a major process/step to most of these early models is the concept of the counter factual (which we'll cover in depth later in this presentation).
+## The Counterfactual
 
-The general concept of the counter factual is it takes our "what if?" questions from implicit to explicit in our data.
+- A major process/step to most of these early models is the concept of the counter factual (which we'll cover in depth later in this presentation).
+
+- The general concept of the counter factual is it takes our "what if?" questions from implicit to explicit in our data.
 
@@
 
-*3. Estimate the statistical parameter (difference/ratio) by substituting the predicted mean outcomes under a = 1 and a = 0*
+*3. Estimate the statistical parameter (difference/ratio) by substituting the predicted mean outcomes under a=1 and a=0*
 
@@
 ```
-
 *Risk Ratio of E(Y|A)*
-
 ```{r ratio}
@@
 ## Updating the Predicted Outcomes Under a = 1 
-
 First we duplicate our original dataframe and then set everyone's treatment to 1.
-
 ```{r create a1 exposure dataframe}
@@
 ```
-
 Now we apply our saved model coefficients (ratio_model) and updated the predicted values using the dataframe where everyone receives treatment E(Y|a=1).
-
 ```{r pr a1}
@@
 ## Updating the Predicted Outcomes Under a = 0
-
 We repeat the same step under the scenario where no one is treated.
-
-```{r create a1 exposure dataframe}
+```{r create a0 exposure dataframe}
 # set everyone to exposed from our original dataframe
@@
 ```
-
 Updating our predicted probability under no treatment scenario E(Y|a=0).
-
 ```{r pr a0}
@@
 ## Estimate the Counterfactual Parameter of Interest
-
 Now that we have a mean value for Y where everyone is treated and no one is treated, we can estimate our risk ratio.
-
 ```{r g forumla ratio}
+# g formula ratio
 mean(pr_a1)/mean(pr_a0)
 ```
-
 Let's compare to our estimate using logistic regression again.
-
 ```{r standard est result}
+# odds ratio (approximates risk ratio due to rare disease assumption)
 exp(ratio_model$coefficients[2])
@@
 
-We get pretty much the same result. That was a lot of steps to get the same answer.
+We get pretty much the same result using the g-formula approach to a standard regression.
+
+## Counterfactual Estimate using IPTW

### 2017-07-12 msm_introduction.Rmd
updated slidy presentation.

@@
 
-- Two general categories that have been around for a decade or so
+- Two general categories of commonly used causal inference methods
 
-1. Structural nested models 
-- Commonly referred to as g-estimation, g-computation, g-formula, g-something
-- The g stands for "general"
+1. Parametric g-formula (sometimes called simple substitution)
 
-2. Marginal structural models 
-- Sometimes called inverse-probability-of-treatment weighting (IPTW)
-
-- Newer approach in recent years are the doubly-robust method which we will touch on later
+2. Marginal structural models (Inverse-probability-of-treatment weighting [IPTW])
 
@@
 
-We get pretty much the same result using the g-formula approach to a standard regression.
+We get pretty much the same result using the g-formula approach to a standard regression. However, we've now made our "what if?" question explicit and testable.
 
 ## Counterfactual Estimate using IPTW
+Another commonly use method is the inverse-probability-of-treatment weight.

### 2017-07-19 msm_introduction.Rmd
updated presentation.

@@
 These models help us answer the marginal effects of a binary exposure and help us deal with assessing the marginal effect of a binary exposure that varies over time, and is also confounded over time (original motivation of Robins).
-These methods can be applied to continous or binary outcomes, but only binary exposures.
+These methods can be applied to continuous or binary outcomes, but only binary exposures.
 
@@
 Judea Pearl and Jamie Robins also did some stuff in the 80's.
-Jamie Robins introduced the g-formula and IPTW which were designed to estimate the marginal effect of a binary treatment that varied over time and was confounded by a time-varying confounder.
+Jamie Robins introduced the g-formula and IPTW which were designed to estimate the marginal effect of a binary treatment that varied over time and was confounded by a time-dependent confounded.
 Jamie Robins and a lot of his students have made some tweaks to the original framework which pretty much brings us to today.
@@
 ## Updating the Predicted Outcomes Under a = 1 
-First we duplicate our original dataframe and then set everyone's treatment to 1.
+First we duplicate our original data frame and then set everyone's treatment to 1.
 ```{r create a1 exposure dataframe}
@@
 ```
-Now we apply our saved model coefficients (ratio_model) and updated the predicted values using the dataframe where everyone receives treatment E(Y|a=1).
+Now we apply our saved model coefficients (ratio_model) and updated the predicted values using the data frame where everyone receives treatment E(Y|a=1).
 ```{r pr a1}
@@
 # g formula ratio
-mean(pr_a1)/mean(pr_a0)
+round(mean(pr_a1)/mean(pr_a0),2)
 ```
@@
 # odds ratio (approximates risk ratio due to rare disease assumption)
-exp(ratio_model$coefficients[2])
+round(exp(ratio_model$coefficients[2]),2)
 ```
 
-We get pretty much the same result using the g-formula approach to a standard regression. However, we've now made our "what if?" question explicit and testable.
+We get pretty much the same result using the g-formula approach to a standard regression.
+The reason for the difference is that I'd normally use some type of Poisson model to estimate a risk ratio, but I used a logistic model to estimate the predicted probability.
+
+However, the important thing is that we've now made our "what if?" question explicit and testable.
 
 ## Counterfactual Estimate using IPTW
-Another commonly use method is the inverse-probability-of-treatment weight.
+Another commonly use method is the inverse-probability-of-treatment weight.This is probably the most common marginal structural models (MSMs) methods used in practice today.
+
+With IPTW estimation, confounding can be thought of as a problem of bias sampling where certain exposure-covariate subgroups are over or under represented compared to an randomized control trail.
+IPTW up-weights under-represented subjects from exposure-covariate strata.
+
+## Counterfactual Estimation using IPTW
+Steps to counter factual estimation are as follows:
+
+*1. Estimate the propensity score Pr(A|W)*
+
+*2. Create weights*
+
+*3. Estimate statistical parameter*
+
+## Estimation of the Propensity of Pr(A)
+
+Much like g-computation, our first step involves a standard regression model.
+However, unlike g-computation, the first step is to estimate the probability of treatment/exposure (A), given any potential covariates/confounders.
+Since we don't have any simulated confounders, we'll just estimate the probability of treatment (which should be pretty close to 0.5 based on our simulation)
+
+```{r iptw step 1}
+prop.table(xtabs(~A, sim_data))
+# intercept only model since we don't have covariates
+prop_mod <- glm(A ~ 1, family = "binomial", data = sim_data)
+```
+
+## Estimating Probability of A = 1 and A = 0
+
+IPTW requires the assumption of positivity or experimental treatment assignment (ETA) where the observed treatment levels vary within confounded strata.
+Positivity violations occur when certain subgroups in a sample rarely or never (or almost always) receive some treatments of interest.
+The resulting sparsity in the data may increase bias with or without an increase in variance and can threaten valid inference.
+
+Before we create the weights, we first need to see if we have an ETA violation.
+Since these are simulated data and there are no factors that affect treatment probability, there won't be.
+The histogram doesn't make sense in this case, so I"m commenting it out.
+
+```{r iptw step 2.1}
+# vector of propensity exposed for each observation
+prob_exposed <- predict(prop_mod, type = "response")
+# vector of propensity not exposed
+prob_unexposed <- 1 - prob_exposed 
+# histogram to check for ETA violation
+#hist(prob_exposed)
+```
+
+## Creating Weights 
+The weight (wt) is simply 1 over the probability of being exposed or unexposed. For those who were exposed, we assign wt as 1/pr(A=1). For unexposed, we assign wt as 1/pr(A=0)
+
+```{r iptw step 2.2}
+# create weights
+wt <- as.numeric(sim_data$A == 1)/prob_exposed + 
+  as.numeric(sim_data$A == 0)/prob_unexposed
+
+# checking to make sure we inverted probabilities correctly
+head(data.frame(sim_data$Y, sim_data$A, 1/prob_exposed, 1/prob_unexposed, wt))
+```
+
+## Estimating the Statistical Parameter
+
+Now that we have the weights, we take all the subjects where A=1, multiply the weight and outcome and divide that by all the subjects where A=0 multiplied by the weights and outcome.
+
+```{r iptw step 3}
+iptw <- mean(as.numeric(sim_data$A==1)*wt*sim_data$Y)/
+  mean(as.numeric(sim_data$A==0)*wt*sim_data$Y)
+# estimate
+round(iptw,2)
+```
+
+I'm not sure this is a counter factual estimate, but it is considered a causal estimate. Note: may revise all slides to focus on "causal inference methods" rather than counter factual.
+
+## What's the Point?
+You've probably thought that this is a lot of work to arrive at the same point estimate that can be estimated with a line of code.
+
+Also, you've probably noticed that we haven't estimated any sort of inferential statistics like confidence intervals.
+Both g-computation and IPTW don't have empirical variance formulas (to my knowledge), therefore confidence intervals need to be estimated by bootstrapping or other sandwich approximations.
+
+There are two scenarios where these 'causal' estimates will differ from standard regression approaches.
+
+1. The estimation of treatment on outcome that is confounded by a time-dependent factor.
+
+2. The estimation of treatment on outcome in the presence of interaction.
+
+## Causal Estimates when A is Under the Influence of a Time-Dependent Confounder
+
+The g-computation and IPTW can be used to give a marginal estimate of the effect of a binary exposure/treatment (A) on an outcome (Y) when A is administered at multiple time points.
+Further, A at time 0 influence a third factor Z, and that factor in turn influences A at time 1.
+Thus, our A and Y association is confounded by a time-dependent factor (Z).
+
+This is a little confusing, so we'll show an example.
+I'm going to use the example from Naimi et al.
+2017, *An introduction to g methods*.
+It's a nicely written paper for those new to these causal inference methods and it also has SAS code for SAS users.
+
+## Simulating Example Shown in Naimi 2017
+The Naimi paper uses the SAS cards statement to provide strata estimates. I've simulated a nearly identical data frame (code hidden due to size).
+
+```{r time dependent Z sim data, echo=F}
+# hiding simulation since it takes up a lot of space
+# i'm sure there is a better way to simulate this, but eh
+# simulating table 1 (kind of)
+set.seed(123)
+strata1 <- tibble(obs = 1:209271) %>%
+  mutate(strata = 1, A0 = 0, Z1 = 0, A1 = 0, 
+    # i'm simulating a little variance around the value provided by Naimi
+    Y = rnorm(n = 209271, mean = 87.29, sd = 10))
+
+strata2 <- tibble(obs = 1:93779) %>% 
+  mutate(strata = 2, A0 = 0, Z1 = 0, A1 = 1, 
+    Y = rnorm(n = 93779, mean = 112.11, sd = 12))
+
+strata3 <- tibble(obs = 1:60654) %>%
+  mutate(strata = 3, A0 = 0, Z1 = 1, A1 = 0, 
+    Y = rnorm(n = 60654, mean = 119.65, sd = 16))
+
+strata4 <- tibble(obs = 1:136293) %>%
+  mutate(strata = 4, A0 = 0, Z1 = 1, A1 = 1, 
+    Y = rnorm(n = 136293, mean = 144.84, sd = 10))
+
+strata5 <- tibble(obs = 1:134781) %>%
+  mutate(strata = 5, A0 = 1, Z1 = 0, A1 = 0, 
+    Y = rnorm(n = 134781, mean = 105.28, sd = 14))
+
+strata6 <- tibble(obs = 1:60789) %>%
+  mutate(strata = 6, A0 = 1, Z1 = 0, A1 = 1, 
+    Y = rnorm(n = 60789, mean = 130.18, sd = 11))
+
+strata7 <- tibble(obs = 1:93903) %>%
+  mutate(strata = 7, A0 = 1, Z1 = 1, A1 = 0, 
+    Y = rnorm(n = 93903, mean = 137.72, sd = 14))
+
+strata8 <- tibble(obs = 1:210527) %>%
+  mutate(strata = 8, A0 = 1, Z1 = 1, A1 = 1, 
+    Y = rnorm(n = 210527, mean = 162.83, sd = 9))
+
+# bind all in to one dataset
+td_sim_data <- rbind(strata1, strata2, strata3, strata4, strata5, strata6, 
+                     strata7, strata8)
+# remove to save room
+rm(strata1, strata2, strata3, strata4, strata5, strata6, strata7, strata8)
+```
+
+Summary statistics for the simulated dataframe that approximate table 1 in Naimi.
+```{r summary td data frame}
+summary_td_sim_data <- td_sim_data %>% group_by(strata) %>% 
+  summarise(A0 = mean(A0), Z1 = mean(Z1), A1 = mean(A1), Y = mean(Y),
+            N = n())
+
+summary_td_sim_data
+```
+
+A benefit of these approaches over standard regression analyses is that they can provide estimates of the marginal effect of A on Y in the presence of interaction with A.

### 2017-07-19 msm_introduction.Rmd
my package failed to install.
need to change something in the msmodelr package.

@@
 
+Standard regression estimate.
+
+```{r time dependent std regression}
+std_td_mod <- glm(Y~A0 + A1 + A0*A1, family = "gaussian", data = td_sim_data)
+
+```
+
+Now that we have our simulated data, we'll replicate the results Naimi et al.
+of marginal (overall) effect of HIV treatment on CD4 count.
+
+To make things easier, let's load some functions I've made for g-computation and IPTW.
+
+```{r}
+# install custom package MSModelR
+# devtools::install_github("RyanGan/MSModelR")
+
+# load library
+library(MSModelR)
+```
+
 A benefit of these approaches over standard regression analyses is that they can provide estimates of the marginal effect of A on Y in the presence of interaction with A.

### 2017-07-19 msm_introduction.Rmd
I can't figure out time-dependent example and may scrap it for now.

@@
 ## Causal Estimates when A is Under the Influence of a Time-Dependent Confounder
+Note I can't figure out how to estimate these values yet so I may cut it out for now.
 
@@
 
-```{r time dependent std regression}
-std_td_mod <- glm(Y~A0 + A1 + A0*A1, family = "gaussian", data = td_sim_data)
+```{r time dependent std regression, echo = F}
+std_est_table <- as_tibble(matrix(NA, nrow = 6, ncol = 2))
+colnames(std_est_table) <- c("model_params","a_estimate")
 
+# generate a vector of formulas to estimate and fill first column
+std_est_table[,1] <- formula_vector <- c("Y~A0+A1+A0*A1", "Y~A0+A1+A0*A1+Z1", 
+  "Y~A0", "Y~A0+Z1", "Y~A1", "Y~A1+Z1")
+
+# for loop to estimate standard regression estimates
+for(i in 1:6){
+  x <- formula_vector[i]
+  mod <- lm(as.formula(x), data=td_sim_data)
+  std_est_table[i,2] <- round(sum(mod$coefficients[
+    # subset only A0, A1, or A0*A1
+    names(mod$coefficients)=="A0" | names(mod$coefficients)=="A1" |
+    names(mod$coefficients)=="A0*A1"]),1) 
+}
+
+knitr::kable(std_est_table, 
+  caption = "Standard regression estimates of average effect of A on Y")
 ```
@@
 # devtools::install_github("RyanGan/MSModelR")
-
 # load library
 library(MSModelR)
+
+test <- g.formula(data = td_sim_data, y = "Y", a = c("A0"), model.family = "gaussian",
+                  q.model = "A0+A1+A0*A1+Z1", estimate = "diff")
+
+iptw(data=td_sim_data, y = "Y", a = "A1", w = "Z1", estimate = "diff")
+
+?iptw
 ```

### 2017-07-31 msm_introduction.Rmd
figured out g-formula est for tv model

@@
 
+Trying to replicate example in sas code.
+```{r}
+y_mod <- glm(Y ~ A1 + Z1 + A0 + A1*A0, data = td_sim_data, family = "gaussian")
+z1_mod <- glm(Z1 ~ A0, data = td_sim_data, family = "binomial"(link="logit"))
+a1_mod <- glm(A1 ~ Z1, data = td_sim_data, family = "binomial"(link="logit"))
+
+# natural course estimate
+nc_mu <- mean(predict(y_mod, newdata = td_sim_data, type = "response"))
+
+# fully exposed
+z1_pr <- mean(predict(z1_mod, newdata = mutate(td_sim_data, A0 = 1), type = "response"))
+# simulate vector of new Z1 values accounting for prior A0
+z1_vec <- ifelse(rbernoulli(n = nrow(td_sim_data), z1_pr)==T,1,0)
+# update a1 
+a1_pr <- mean(predict(a1_mod, newdata = as.tibble(z1_vec), type = "response"))
+# simulate new a1 given z1 under a0 estimate
+a1_vec <- ifelse(rbernoulli(n = nrow(td_sim_data), a1_pr)==T,1,0)
+
+all_exposed <- mean(predict(y_mod, newdata = mutate(td_sim_data, A1 = 1, A0 = 1, Z1 = z1_vec),
+                type = "response"))
+
+head(mutate(td_sim_data, A1 = 1, A0 = 1))
+
+MSModelR::g.formula
+```
+
 To make things easier, let's load some functions I've made for g-computation and IPTW.

### 2017-07-31 msm_introduction.Rmd
figured out how to reproduce table in naimi.

--- 2017-07-19 msm_introduction.Rmd
@@
 
+Trying to replicate example in sas code.
+```{r}
+y_mod <- glm(Y ~ A1 + Z1 + A0 + A1*A0, data = td_sim_data, family = "gaussian")
+z1_mod <- glm(Z1 ~ A0, data = td_sim_data, family = "binomial"(link="logit"))
+a1_mod <- glm(A1 ~ Z1, data = td_sim_data, family = "binomial"(link="logit"))
+
+# natural course estimate ----
+nc_mu <- mean(predict(y_mod, newdata = td_sim_data, type = "response"))
+
+# fully exposed ----
+z1_pr <- mean(predict(z1_mod, newdata = mutate(td_sim_data, A0 = 1), type = "response"))
+# simulate vector of new Z1 values accounting for prior A0
+z1_vec <- ifelse(rbernoulli(n = nrow(td_sim_data), z1_pr)==T,1,0)
+
+# update a1 
+a1_pr <- mean(predict(a1_mod, newdata = mutate(td_sim_data, Z1 = z1_vec), 
+                      type = "response"))
+# simulate new a1 given z1 under a0 estimate
+a1_vec <- ifelse(rbernoulli(n = nrow(td_sim_data), a1_pr)==T,1,0)
+# update predicted Y where everyone is exposed at time 0 and 1, and accounting for Z1 | A0 =1 
+all_exposed <- mean(predict(y_mod, newdata = mutate(td_sim_data, A1 = 1, A0 = 1, Z1 = z1_vec),
+                type = "response"))
+all_exposed
+
+# unexposed
+z1_pr <- mean(predict(z1_mod, newdata = mutate(td_sim_data, A0 = 0), type = "response"))
+z1_vec <- ifelse(rbernoulli(n = nrow(td_sim_data), z1_pr)==T,1,0)
+all_unexposed <- mean(predict(y_mod, newdata = mutate(td_sim_data, A1 = 0, A0 = 0, Z1 = z1_vec),
+                type = "response"))
+all_unexposed
+# diff
+all_exposed - all_unexposed
+
+```
+
 To make things easier, let's load some functions I've made for g-computation and IPTW.

### 2017-08-01 msm_introduction.Rmd
updated some stuff

@@
 
-Trying to replicate example in sas code.
+Trying to replicate simple substitution example in sas code.
 ```{r}
+# g formula
 y_mod <- glm(Y ~ A1 + Z1 + A0 + A1*A0, data = td_sim_data, family = "gaussian")
@@
 # fully exposed ----
-z1_pr <- mean(predict(z1_mod, newdata = mutate(td_sim_data, A0 = 1), type = "response"))
+z1_pr <- mean(predict(z1_mod, newdata = mutate(td_sim_data, A0 = 1), 
+                      type = "response"))
 # simulate vector of new Z1 values accounting for prior A0
@@
 # update predicted Y where everyone is exposed at time 0 and 1, and accounting for Z1 | A0 =1 
-all_exposed <- mean(predict(y_mod, newdata = mutate(td_sim_data, A1 = 1, A0 = 1, Z1 = z1_vec),
-                type = "response"))
+all_exposed <- mean(predict(y_mod, newdata = mutate(td_sim_data, 
+                A1 = 1, A0 = 1, Z1 = z1_vec), type = "response"))
 all_exposed
@@
 # unexposed
-z1_pr <- mean(predict(z1_mod, newdata = mutate(td_sim_data, A0 = 0), type = "response"))
+z1_pr <- mean(predict(z1_mod, newdata = mutate(td_sim_data, A0 = 0), 
+          type = "response"))
 z1_vec <- ifelse(rbernoulli(n = nrow(td_sim_data), z1_pr)==T,1,0)
-all_unexposed <- mean(predict(y_mod, newdata = mutate(td_sim_data, A1 = 0, A0 = 0, Z1 = z1_vec),
-                type = "response"))
+all_unexposed <- mean(predict(y_mod, newdata = mutate(td_sim_data, 
+                  A1 = 0, A0 = 0, Z1 = z1_vec), type = "response"))
 all_unexposed
@@
 all_exposed - all_unexposed
+
+# creating a formula. will need dplyr
+g.formula.td <- function(data, y, a0, a1, z1, y.model, y.family,
+  a1.model, a1.family, z1.model, z1.family, estimate = "diff"){
+
+  un_exp_df <- exp_df <- data
+  un_exp_df[,c(a0,a1)] <- 0
+  exp_df[, c(a0,a1)] <- 1
+
+  # define y.model
+  q_mod <- glm(as.formula(paste(y, y.model, sep = "~")), data = data, 
+    family = y.family)
+  # define z1.model
+  z1_mod <- glm(as.formula(paste(z1, z1.model, sep = "~")), data = data, 
+    family = z1.family)
+  # define a1.model
+  a1_mod <- glm(as.formula(paste(a1, a1.model, sep = "~")), data = data, 
+              family = a1.family) 
+  # updated probability of z1 given a0 = 1
+  z1_exp_pr <- mean(predict(z1_mod, newdata = exp_df, type = "response"))
+  # simulate vector of new Z1 values accounting for prior A0 where everyone is exposed
+  z1_exp_vec <- as.numeric(ifelse(rbinom(n = nrow(data),size = 1 , z1_exp_pr)==T,1,0))
+  # update z value under a0=1 scenario
+  exp_df[,z1] <- z1_exp_vec
+  # update y estimate as if everyone was exposed
+  y_exp <- mean(predict(y_mod,newdata=exp_df,type = "response"))
+
+  # updated probability of z1 given a0 = 0
+  z1_unexp_pr <- mean(predict(z1_mod, newdata = un_exp_df, type = "response"))
+  # simualte vector of new Z1 values accounting for prior A0 where everyone is unexposed
+  z1_unexp_vec <- as.numeric(ifelse(rbinom(n = nrow(data), size = 1, z1_unexp_pr)==T,1,0))
+  un_exp_df[,z1] <- z1_unexp_vec
+  # udpate y estimates as if everyone was unexposed
+  y_unexp <- mean(predict(y_mod,newdata=un_exp_df,type = "response"))
+
+  if(estimate == "diff"){ 
+    msm_estimate <- y_exp - y_unexp
+    }
+  if(estimate == "ratio"){
+    msm_estimate <- y_exp/y_unexp
+  }
+  return(msm_estimate)
+  }
+
+g.formula.td(data = td_sim_data, y = "Y", a0 = "A0", a1= "A1", z1 = "Z1",
+  y.model = "A0+A1+Z1+A0*A1", y.family = "gaussian", a1.model = "Z1", 
+  a1.family = "binomial", z1.model = "A0", z1.family = "binomial", estimate = "diff")
 

### 2017-08-01 msm_introduction.Rmd
updated msm

@@
 
+```{r iptw td}
+MSModelR::iptw
+
+# define propensity models (intercept only model without covariates)
+pr_a0_mod <- glm(A0 ~ 1, data = td_sim_data, family="binomial")
+# probability of being treated at time 1
+pr_a0 <- mean(predict(pr_a0_mod, type = "response"))
+# pr a1 | stuff
+
+```
 To make things easier, let's load some functions I've made for g-computation and IPTW.

### 2017-08-02 msm_introduction.Rmd
figured out iptw to td model.

--- 2017-08-01 msm_introduction.Rmd
@@
 
-If you've been to Society for Epidemiologic Research or any epi methods conference, you've probably heard something about "causal inference".
+Causal inference methods are very popular (particularly with methods epidemiologists)
 
-You've also probably heard a lot of gibberish about g-estimation, g-computation, g-formula, g-something, inverse-probability-of-treatment weighting, marginal structural models, doubly-robust, targeted maximum likelihood, etc.
+Terms associated with thesee methods include: DAGs, g-estimation, g-computation, g-formula, g-something, inverse-probability-of-treatment weighting, marginal structural models, doubly-robust estimation, targeted maximum likelihood, etc.
 
-And like me, you probably have no idea what any of those are, what they are used for, or even how to use them.
-But they are always referred to as being used in "causal inference", so these fancy-pants models must be special and the only way to find causal associations.
+This presentation provides an introduction to two of the more common of these methods.
 
-This presentation provides a general overview of things I've learned about these methods.
-I'm still confused about a lot of this topic area (like how is the g-formula different from g-estimation?), but hopefully this will server as a primer.
+## What's the big deal about causal inference methods?
+
+ -
 
@@
 
+```{r iptw td}
+MSModelR::iptw
+
+# define propensity models (intercept only model without covariates)
+pr_a0_mod <- glm(A0 ~ 1, data = td_sim_data, family="binomial")
+# probability of being treated at time 1
+pr_a0 <- predict(pr_a0_mod, type = "response")
+# define propensity a1 model
+pr_a1_mod <- glm(A1 ~ 1, data = td_sim_data, family = "binomial")
+# probability of being treated at time 2
+pr_a1 <- predict(pr_a1_mod, type = "response")
+# probability of a1 given z1
+pr_a1_z_mod <- glm(A1 ~ Z1, data = td_sim_data, family = "binomial")
+# probability of a1 given z1
+pr_a1_z <- predict(pr_a1_z_mod, type = "response")
+
+# create weights in dataframe
+td_sim_data <- td_sim_data %>% 
+  mutate(wt = ifelse(A0 == 1 & A1 == 1, (pr_a0/pr_a0)*(pr_a1/pr_a1_z),
+    ifelse(A0 == 1 & A1 == 0, (pr_a0/pr_a0)*((1-pr_a1)/(1-pr_a1_z)),
+    ifelse(A0 == 0 & A1 == 1, ((1-pr_a0)/(1-pr_a0))*(pr_a1/pr_a1_z),
+    ifelse(A0 == 0 & A1 == 0, ((1-pr_a0)/(1-pr_a0))*((1-pr_a1)/(1-pr_a1_z)),NA
+    )))))
+
+# table of summarized weights
+wt_table <- td_sim_data %>% 
+  group_by(strata) %>% 
+  summarise(A0 = mean(A0), Z1 = mean(Z1), A1 = mean(A1), Y = mean(Y), n = n(),
+            wt = mean(wt), pseudo_n = n*wt)
+
+wt_table
+
+# weighted model
+iptw_mod <- glm(Y ~ A0 + A1 + A0*A1, data = td_sim_data, weights = wt,
+                family = "gaussian")
+
+summary(iptw_mod)
+
+sum(iptw_mod$coefficients[names(iptw_mod$coefficients)=="A0" | 
+  names(iptw_mod$coefficients)=="A1" | names(iptw_mod$coefficients)=="A0*A1"],1)
+
+```
 To make things easier, let's load some functions I've made for g-computation and IPTW.

### 2017-08-02 msm_introduction.Rmd
updated presentation

@@
 ---
-title: "Introduction to Causal Inference Models"
+title: "Introduction to Marginal Structural Models"
 author: "Ryan Gan"
 date: "7/10/2017"
-output: slidy_presentation
+output: 
+  slidy_presentation:
+    incremental: true
 ---
@@
 
-## "Causal Inference"
+## Causal inference
 
-Causal inference methods are very popular (particularly with methods epidemiologists)
+- Causal inference methods are growing in popularity (particularly among the epidemiology methodologists)
 
-Terms associated with thesee methods include: DAGs, g-estimation, g-computation, g-formula, g-something, inverse-probability-of-treatment weighting, marginal structural models, doubly-robust estimation, targeted maximum likelihood, etc.
+- Terms associated with thesee methods include: DAGs, counterfactual, potential outcomes, g-estimation, g-computation, g-formula, g-something, inverse-probability-of-treatment weighting, marginal structural models, doubly-robust estimation, targeted maximum likelihood, etc.
 
-This presentation provides an introduction to two of the more common of these methods.
+- Jargon around this field makes these methods intimidating
+
+- This presentation will focus on marginal structural models (MSMs) and provide an introduction to two of the more common of these methods.
 
@@
 
- -
+- Another tool/skillset to help evaluate causal relationships
 
-## Causal Inference Models
+- Emphasis mostly placed on the analysis of data
 
-These models help us answer the marginal effects of a binary exposure and help us deal with assessing the marginal effect of a binary exposure that varies over time, and is also confounded over time (original motivation of Robins).
-These methods can be applied to continuous or binary outcomes, but only binary exposures.
+- Causal inference language is different from statistical language (i.e. X affecting Y vs. X correlated with Y) although we may run the exact same statistical test to assess
 
-- Two general categories of commonly used causal inference methods
+## Three main methods you'll hear about
+
+1.
+Directed Acyclic Graphs (DAGs): help to understand potential relationships between variables (i.e.
+potential collider and confounding bias)and how data may have been generated (i.e.
+temporal sequence exposure affects outcome)
+
+2. Counterfactual/potential outcomes: makes explicit the "what if?" scenario
+
+3.
+MSMs provide consistent estimates of differences or ratios of average (marginal) potential outcomes with less restrictions compared to standard regression methods (we'll return to this when we talk about time-dependent confounding)
+
+- Keep in mind your training in standard regression and this will make more sense!
+
+## What is a counter factual?
+
+- Randomized control trial: randomly assign exposure and no exposure to two groups and follow them over time to see which group is more likely to develop the outcome
+
+- What is implicit in this design is that the 'no exposure' group represents what would have happened to the 'exposure' group
+
+- A missing data problem: a subject's exposure and outcome can be observed only once
+
+- The counter factual solves our missing data problem and takes our "what if?" questions from implicit to explicit in our analysis of data
+
+- You will also see this called potential outcomes framework in some circles (economists)
+
+## Marginal Structural Models (MSMs)
+
+- Uses the pricinipal of the counter factual and allows us to estimate an unbiased and consistent effect,under a less-restrictive set of assumptions:
+
+1. Counterfactual consistency
+
+2. Exchangebility
+
+3. Positivity
+
+## Counterfactual consistency
+
+- Allows us to equate observed outcomes for a given exposure value to the potential outcomes that would be observed under the same exposure value.
+
+- This assumption depends on exposure assignment mechanism (i.e. RCT)
+
+## Exchangeability
+
+- Potential outcome under exposure a~0~ and a~1~ are independent of the observed exposures A~0~ and A~1~
+
+- Can be tested emprically
+
+- Easier to understand as being able to exchange exposure groups and observing the same distribution of the outcome
+
+- Conditional exchangeability happens when there is no unmeasured confounding
+
+## Positivity
+
+- Probability of exposure is not 0 or 1
+
+- Can be emprically evaluated to make sure there are exposed and unexposed individuals within each confounder strata and prior exposure levels
+
+## Two commonly used MSMs
+
+While there are more classes of MSMs we'll cover the most basic two
 
@@
 
-2. Marginal structural models (Inverse-probability-of-treatment weighting [IPTW])
+2. Inverse-probability-of-treatment weighting (IPTW)
 
-## A Brief History of Causal Models
+## Simulated data
 
-A long time ago (1920s), in a country far far away (Poland), Jerzy Neyman proposed a class of statistical models that would latter be referred to as counter factual/potential outcomes models for randomized trials.
-Donald Rubin extended this framework to observational data and coined the term potential outcomes in maybe the 1970s?.
-Judea Pearl and Jamie Robins also did some stuff in the 80's.
-Jamie Robins introduced the g-formula and IPTW which were designed to estimate the marginal effect of a binary treatment that varied over time and was confounded by a time-dependent confounded.
-Jamie Robins and a lot of his students have made some tweaks to the original framework which pretty much brings us to today.
+I've simulated a simple example data frame with a known relationship to show you how to implement these two methods
 
-## The Counterfactual
-
-- A major process/step to most of these early models is the concept of the counter factual (which we'll cover in depth later in this presentation).
-
-- The general concept of the counter factual is it takes our "what if?" questions from implicit to explicit in our data.
-
-## Counterfactuals: Making Explicit Causal Claims
-
-- The randomized control trial (RCT): Our gold standard for assessing causal relationships.
-The basic study design has the investigators randomly assign a binary treatment to a population and follow the population over time to identify incident outcomes.
-Any potential confounding variables are distributed evenly among our treatment group, which creates d-separation/independence/no association between our confounding variables and our treatment.
-With this study design, we can estimate causal and unbiased effects between treatment and outcomes.
-
-- What is implicit in this design is that the 'no' treatment group represents what would have happened to the 'yes' treatment group had they not been given the treatment.
-
-- Counter factual estimation makes this implicit assumption explicit.
-
-## Counterfactual Estimation is a Problem of Missing Data
-
-- The problem with any epidemiologic study design is that a subject's exposure (denoted as A) and outcome (denoted as Y) can be observed only once.
-
-```{r simple simulation}
+Simulate code is hidden, but here are the first 6 observations and a 2x2 table
+```{r simple simulation, echo=FALSE}
 # set sample size and seed 
@@
 head(sim_data)
+#2x2 table
+xtabs(~A + Y, sim_data)
 ```
 
-- For example, our first subject has been treated and has not developed the outcome.
+## MSM using g-computation
 
-## Counterfactual Estimate using G-Computation
+- G-computation is a way in which we can estimate the counter factual state where everyone is treated/exposed (denoted by a = 1) and no one is treated/exposed (denoted by a = 0) and is implemented in a 3-step process:
 
-- G-computation is a way in which we can estimate the counter factual state where everyone is treated/exposed (denoted by a = 1) and no one is treated/exposed (denoted by a = 0)
+1. Estimation of the conditional mean outcome E(Y|A)
 
-- G-computation is implemented in a 3-step process:
+2. Obtaining predicted potential outcomes under a = 1 and a = 0
 
-*1. Estimation of the conditional mean outcome E(Y|A)*
+3. Estimate the statistical parameter (difference/ratio) by substituting the predicted mean outcomes under a=1 and a=0
 
-*2. Obtaining predicted outcomes under a = 1 and a = 0*
-
-*3. Estimate the statistical parameter (difference/ratio) by substituting the predicted mean outcomes under a=1 and a=0*
-
-## Estimation of the Conditional Mean Outcome E(Y|A)
+## Estimation of the conditional mean outcome E(Y|A)
 
@@
 
-*2x2 Table*
-```{r 2x2}
-xtabs(~A + Y, sim_data)
-```
 *Risk Ratio of E(Y|A)*
-```{r ratio}
+```{r risk ratio}
 # standard ratio model
@@
 # ratio estimate
-exp(ratio_model$coefficients[2])
+round(exp(ratio_model$coefficients[2]),2)
 ```
-The treated group is 40% less likely to develop the outcome compared to the untreated group.
+The treated group is 40% less likely to develop the outcome compared to the untreated group
 
-## Updating the Predicted Outcomes Under a = 1 
-First we duplicate our original data frame and then set everyone's treatment to 1.
+## Updating the predicted outcomes under a=1
+
+First we duplicate our original data frame and then set everyone's exposure to 1
+
 ```{r create a1 exposure dataframe}
@@
 ```
-Now we apply our saved model coefficients (ratio_model) and updated the predicted values using the data frame where everyone receives treatment E(Y|a=1).
+
+Now we apply our saved model coefficients (ratio_model) and updated the predicted values using the data frame where everyone receives exposure E(Y|a=1)
+
 ```{r pr a1}
@@
 
-## Updating the Predicted Outcomes Under a = 0
+## Updating the predicted outcomes under a=0
+
 We repeat the same step under the scenario where no one is treated.
@@
 ```
-Updating our predicted probability under no treatment scenario E(Y|a=0).
+Updating our predicted probability under no exposure scenario E(Y|a=0).
 ```{r pr a0}
@@
 
-## Estimate the Counterfactual Parameter of Interest
+## Estimate the counterfactual parameter of interest
+
 Now that we have a mean value for Y where everyone is treated and no one is treated, we can estimate our risk ratio.
@@
 We get pretty much the same result using the g-formula approach to a standard regression.
-The reason for the difference is that I'd normally use some type of Poisson model to estimate a risk ratio, but I used a logistic model to estimate the predicted probability.
 
@@
 
-## Counterfactual Estimate using IPTW
-Another commonly use method is the inverse-probability-of-treatment weight.This is probably the most common marginal structural models (MSMs) methods used in practice today.
+## MSM via IPTW
+
+Another commonly use method is the inverse-probability-of-treatment weight.This is probably the most common marginal structural models (MSMs) methods used in practice today because it's easier to implement
 
@@
 
-## Counterfactual Estimation using IPTW
-Steps to counter factual estimation are as follows:
+## IPTW steps
 
-*1. Estimate the propensity score Pr(A|W)*
+1. Estimate the propensity score Pr(A|W)
 
-*2. Create weights*
+2. Create weights
 
-*3. Estimate statistical parameter*
+3. Estimate statistical parameter
 
-## Estimation of the Propensity of Pr(A)
+## Estimation of the propensity of Pr(A)
 
-Much like g-computation, our first step involves a standard regression model.
-However, unlike g-computation, the first step is to estimate the probability of treatment/exposure (A), given any potential covariates/confounders.
-Since we don't have any simulated confounders, we'll just estimate the probability of treatment (which should be pretty close to 0.5 based on our simulation)
+Much like g-computation, our first step involves a standard regression model. However, unlike g-computation, the first step is to estimate the probability of treatment/exposure (A).
 
 ```{r iptw step 1}
-prop.table(xtabs(~A, sim_data))
 # intercept only model since we don't have covariates
 prop_mod <- glm(A ~ 1, family = "binomial", data = sim_data)
+# summary of model
 ```
 
-## Estimating Probability of A = 1 and A = 0
+Easy enough
 
-IPTW requires the assumption of positivity or experimental treatment assignment (ETA) where the observed treatment levels vary within confounded strata.
-Positivity violations occur when certain subgroups in a sample rarely or never (or almost always) receive some treatments of interest.
-The resulting sparsity in the data may increase bias with or without an increase in variance and can threaten valid inference.
+## Estimating probability of a=1 and a=0
 
-Before we create the weights, we first need to see if we have an ETA violation.
-Since these are simulated data and there are no factors that affect treatment probability, there won't be.
-The histogram doesn't make sense in this case, so I"m commenting it out.
+- For each observation, we estimate the probability of being exposed and simply take 1 minus that probability to calculate the probability of not being exposed
+
+- Before we create the weights, we first need to see if we have positivity, or experimental treatment assignment (ETA),violations (since these are simulated data and there are no factors that affect exposure probability, there won't be)
+
+- Assessing ETA vilations is easy enough to do by looking at a distribution of the predicted probabilities of A
 
@@
 #hist(prob_exposed)
+head(prob_exposed)
 ```
 
-## Creating Weights 
-The weight (wt) is simply 1 over the probability of being exposed or unexposed. For those who were exposed, we assign wt as 1/pr(A=1). For unexposed, we assign wt as 1/pr(A=0)
+## Creating weights
+
+- The weight (wt) is simply 1 over the probability of being exposed or unexposed. For those who were exposed, we assign wt as 1/pr(a=1). For unexposed, we assign wt as 1/pr(a=0)
 
@@
 
-## Estimating the Statistical Parameter
+## Estimating the statistical parameter
 
@@
 ```{r iptw step 3}
-iptw <- mean(as.numeric(sim_data$A==1)*wt*sim_data$Y)/
+# ratio estimate
+iptw_ratio <- mean(as.numeric(sim_data$A==1)*wt*sim_data$Y)/
   mean(as.numeric(sim_data$A==0)*wt*sim_data$Y)
 # estimate
-round(iptw,2)
+round(iptw_ratio,2)
 ```
 
-I'm not sure this is a counter factual estimate, but it is considered a causal estimate. Note: may revise all slides to focus on "causal inference methods" rather than counter factual.
+## What's the point?
 
-## What's the Point?
-You've probably thought that this is a lot of work to arrive at the same point estimate that can be estimated with a line of code.
+- This is a lot of work to arrive at the same point estimate that can be estimated with a line of code
 
-Also, you've probably noticed that we haven't estimated any sort of inferential statistics like confidence intervals.
-Both g-computation and IPTW don't have empirical variance formulas (to my knowledge), therefore confidence intervals need to be estimated by bootstrapping or other sandwich approximations.
+- We haven't estimated any sort of inferential statistics like confidence intervals.
 
-There are two scenarios where these 'causal' estimates will differ from standard regression approaches.
+- G-computation and IPTW don't have empirical variance formulas (to my knowledge), therefore confidence intervals need to be estimated by bootstrapping or other sandwich approximations.
 
-1. The estimation of treatment on outcome that is confounded by a time-dependent factor.
+- There are two scenarios where these 'causal' estimates from MSMs will differ from standard regression approaches and where they are the most appropriate statistical approach
 
-2. The estimation of treatment on outcome in the presence of interaction.
+## MSMs shine when...
 
-## Causal Estimates when A is Under the Influence of a Time-Dependent Confounder
-Note I can't figure out how to estimate these values yet so I may cut it out for now.
+1. The estimation of exposure on outcome is confounded by a time-dependent factor
 
-The g-computation and IPTW can be used to give a marginal estimate of the effect of a binary exposure/treatment (A) on an outcome (Y) when A is administered at multiple time points.
-Further, A at time 0 influence a third factor Z, and that factor in turn influences A at time 1.
-Thus, our A and Y association is confounded by a time-dependent factor (Z).
+2. The estimation of exposure on outcome in the presence of interaction.
 
-This is a little confusing, so we'll show an example.
-I'm going to use the example from Naimi et al.
+## MSM and a time-dependent confounder
+
+- MSMs can help us get unbiased average/marginal estimates of exposure/treatment (A) on an outcome (Y) when A is administered at multiple time points and when a third factor (Z) is effected by A at time 0 and affects A at time 1
+
+- I'm going to use the example from Naimi et al.
 2017, *An introduction to g methods*.
-It's a nicely written paper for those new to these causal inference methods and it also has SAS code for SAS users.
+It's a nicely written paper for those new to these causal inference methods and it also has SAS code for SAS users
 
-## Simulating Example Shown in Naimi 2017
+## Causal structure of our example
+
+![](naimi_fig.png)
+
+## Simulating example Naimi data
+
 The Naimi paper uses the SAS cards statement to provide strata estimates. I've simulated a nearly identical data frame (code hidden due to size).
@@
 
-Summary statistics for the simulated dataframe that approximate table 1 in Naimi.
-```{r summary td data frame}
+```{r summary td data frame, echo=F}
 summary_td_sim_data <- td_sim_data %>% group_by(strata) %>% 
@@
 
-summary_td_sim_data
+knitr::kable(summary_td_sim_data, 
+  caption = "Simulated average outcome values by strata of treatment (A) and confounder (Z)")
 ```
 
-Standard regression estimate.
+## Standard estimates of the average effect of A on Y
+
+For reference, the 'true' average effect of A0 and A1 on Y is an increase of 50 CD4 count
 
@@
 
+## G-formula accounting for time-dependent confounder
+
 Now that we have our simulated data, we'll replicate the results Naimi et al.
-of marginal (overall) effect of HIV treatment on CD4 count.
+of marginal (overall) effect of HIV treatment on CD4 count
 
-Trying to replicate simple substitution example in sas code.
-```{r}
+```{r, eval = F}
 # g formula
@@
 
-```{r iptw td}
+```{r iptw td, eval = F}
 MSModelR::iptw
@@
 
-```{r}
+```{r, eval=F}
 # install custom package MSModelR

### 2017-08-03 msm_introduction.Rmd
updated.

@@
 
+- In the example, we want to know the average effect of a binary treatment for HIV (A) on continous CD4 counts (Y).
+Treatment is administered at multiple time points, with the only other covariate being a binary cutoff of HIV viral load (Z), where Z = 1 for viral load > 200 copies/ml.
+Further, viral load influences the decision to treat at that time point.
+
 ## Causal structure of our example
@@
 ![](naimi_fig.png)
+
+## Testable research questions
+
+- There are a couple that come to mind:
+
+1. What is the average level of CD4 counts in those that received treatment at time 0 compared to those never treated?
 
@@
 ## Standard estimates of the average effect of A on Y
+
+We'd like to know the average effect of treatment (at both time points)
 

### 2017-08-03 msm_introduction.Rmd
Finished msm presentation.

@@
 
-- Causal inference methods are growing in popularity (particularly among the epidemiology methodologists)
+- Causal inference methods are growing in popularity (particularly among the epidemiology methodologists).
 
-- Terms associated with thesee methods include: DAGs, counterfactual, potential outcomes, g-estimation, g-computation, g-formula, g-something, inverse-probability-of-treatment weighting, marginal structural models, doubly-robust estimation, targeted maximum likelihood, etc.
+- Terms associated with these methods include: DAGs, counter factual, potential outcomes, g-estimation, g-computation, g-formula, g-something, inverse-probability-of-treatment weighting, marginal structural models, doubly-robust estimation, targeted maximum likelihood, etc.
 
-- Jargon around this field makes these methods intimidating
+- Jargon around this field makes these methods intimidating.
 
@@
 
-- Another tool/skillset to help evaluate causal relationships
+- Another tool/skill set to help evaluate causal relationships.
 
-- Emphasis mostly placed on the analysis of data
+- Emphasis mostly placed on the analysis of data.
 
-- Causal inference language is different from statistical language (i.e. X affecting Y vs. X correlated with Y) although we may run the exact same statistical test to assess
+- Causal inference language is different from statistical language (i.e. X affecting Y vs. X correlated with Y) although we may run the exact same statistical test to assess.
 
-## Three main methods you'll hear about
+## Three terms you'll frequently encounter
 
@@
 potential collider and confounding bias)and how data may have been generated (i.e.
-temporal sequence exposure affects outcome)
+temporal sequence exposure affects outcome).
 
-2. Counterfactual/potential outcomes: makes explicit the "what if?" scenario
+2. Counter factual/potential outcomes: makes explicit the "what if?" scenario.
 
 3.
-MSMs provide consistent estimates of differences or ratios of average (marginal) potential outcomes with less restrictions compared to standard regression methods (we'll return to this when we talk about time-dependent confounding)
-
-- Keep in mind your training in standard regression and this will make more sense!
+MSMs provide consistent estimates of differences or ratios of average (marginal) potential outcomes with less restrictions compared to standard regression methods (we'll return to this when we talk about time-dependent confounding).
 
@@
 
-- Randomized control trial: randomly assign exposure and no exposure to two groups and follow them over time to see which group is more likely to develop the outcome
+- Randomized control trial: randomly assign exposure and no exposure to two groups and follow them over time to see which group is more likely to develop the outcome.
 
-- What is implicit in this design is that the 'no exposure' group represents what would have happened to the 'exposure' group
+- What is implicit in this design is that the 'no exposure' group represents what would have happened to the 'exposure' group.
 
-- A missing data problem: a subject's exposure and outcome can be observed only once
+- A missing data problem: a subject's exposure and outcome can be observed only once.
 
-- The counter factual solves our missing data problem and takes our "what if?" questions from implicit to explicit in our analysis of data
-
-- You will also see this called potential outcomes framework in some circles (economists)
+- The counter factual solves our missing data problem and takes our "what if?" questions from implicit to explicit in our analysis of data.
 
@@
 
-- Uses the pricinipal of the counter factual and allows us to estimate an unbiased and consistent effect,under a less-restrictive set of assumptions:
+- Uses the principal of the counter factual and allows us to estimate an unbiased and consistent effect,under a less-restrictive set of assumptions:
 
-1. Counterfactual consistency
+1. Counter factual consistency
 
-2. Exchangebility
+2. Exchangeability
 
@@
 
-- This assumption depends on exposure assignment mechanism (i.e. RCT)
+- This assumption depends on exposure assignment mechanism (i.e. RCT).
 
@@
 
-- Potential outcome under exposure a~0~ and a~1~ are independent of the observed exposures A~0~ and A~1~
+- Potential outcome under exposure a~0~ and a~1~ are independent of the observed exposures A~0~ and A~1~.
 
-- Can be tested emprically
+- Can be tested empirically.
 
-- Easier to understand as being able to exchange exposure groups and observing the same distribution of the outcome
+- Easier to understand as being able to exchange exposure groups and observing the same distribution of the outcome.
 
-- Conditional exchangeability happens when there is no unmeasured confounding
+- Conditional exchangeability happens when there is no unmeasured confounding.
 
@@
 
-- Probability of exposure is not 0 or 1
+- Probability of exposure is not 0 or 1.
 
-- Can be emprically evaluated to make sure there are exposed and unexposed individuals within each confounder strata and prior exposure levels
+- Can be empirically evaluated to make sure there are exposed and unexposed individuals within each confounder strata and prior exposure levels.
 
@@
 
-I've simulated a simple example data frame with a known relationship to show you how to implement these two methods
+I've simulated a simple example data frame with a known relationship to show you how to implement these two methods.
 
-Simulate code is hidden, but here are the first 6 observations and a 2x2 table
+Simulate code is hidden, but here are the first 6 observations and a 2x2 table.
 ```{r simple simulation, echo=FALSE}
@@
 
-1. Estimation of the conditional mean outcome E(Y|A)
+1. Estimation of the conditional mean outcome E(Y|A).
 
-2. Obtaining predicted potential outcomes under a = 1 and a = 0
+2. Obtaining predicted potential outcomes under a = 1 and a = 0.
 
-3. Estimate the statistical parameter (difference/ratio) by substituting the predicted mean outcomes under a=1 and a=0
+3. Estimate the statistical parameter (difference/ratio) by substituting the predicted mean outcomes under a=1 and a=0.
 
@@
 
-*Risk Ratio of E(Y|A)*
 ```{r risk ratio}
@@
 ```
-The treated group is 40% less likely to develop the outcome compared to the untreated group
+The treated group is 40% less likely to develop the outcome compared to the untreated group.
 
@@
 
-First we duplicate our original data frame and then set everyone's exposure to 1
-
+First we duplicate our original data frame and then set everyone's exposure to 1.
 ```{r create a1 exposure dataframe}
@@
 ```
-
-Now we apply our saved model coefficients (ratio_model) and updated the predicted values using the data frame where everyone receives exposure E(Y|a=1)
-
+Now we apply our saved model coefficients (ratio_model) and updated the predicted values using the data frame where everyone receives exposure E(Y|a=1).
 ```{r pr a1}
@@
 
-We get pretty much the same result using the g-formula approach to a standard regression.
-
-However, the important thing is that we've now made our "what if?" question explicit and testable.
+We get pretty much the same result using the g-formula approach to a standard regression. However, the important thing is that we've now made our "what if?" question explicit and testable.
 
@@
 
-Another commonly use method is the inverse-probability-of-treatment weight.This is probably the most common marginal structural models (MSMs) methods used in practice today because it's easier to implement
+Another commonly use method is the inverse-probability-of-treatment weight.This is probably the most common marginal structural models (MSMs) methods used in practice today because it's easier to implement.
 
@@
 
-Easy enough
+Easy enough.
 
@@
 
-- For each observation, we estimate the probability of being exposed and simply take 1 minus that probability to calculate the probability of not being exposed
+- For each observation, we estimate the probability of being exposed and simply take 1 minus that probability to calculate the probability of not being exposed.
 
-- Before we create the weights, we first need to see if we have positivity, or experimental treatment assignment (ETA),violations (since these are simulated data and there are no factors that affect exposure probability, there won't be)
+- Before we create the weights, we first need to see if we have positivity, or experimental treatment assignment (ETA),violations (since these are simulated data and there are no factors that affect exposure probability, there won't be).
 
-- Assessing ETA vilations is easy enough to do by looking at a distribution of the predicted probabilities of A
+- Assessing ETA violations is easy enough to do by looking at a distribution of the predicted probabilities of A.
 
@@
 prob_unexposed <- 1 - prob_exposed 
-# histogram to check for ETA violation
-#hist(prob_exposed)
-head(prob_exposed)
 ```
@@
 
-- The weight (wt) is simply 1 over the probability of being exposed or unexposed. For those who were exposed, we assign wt as 1/pr(a=1). For unexposed, we assign wt as 1/pr(a=0)
+- The weight (wt) is simply 1 over the probability of being exposed or unexposed. For those who were exposed, we assign wt as 1/pr(a=1). For unexposed, we assign wt as 1/pr(a=0).
 
@@
 
-- This is a lot of work to arrive at the same point estimate that can be estimated with a line of code
+- This is a lot of work to arrive at the same point estimate that can be estimated with a line of code.
 
@@
 
-## MSMs shine when...
+## When an MSM will provide a better estimate
 
-1. The estimation of exposure on outcome is confounded by a time-dependent factor
+1. The estimation of exposure on outcome is confounded by a time-dependent factor.
 
@@
 
-- MSMs can help us get unbiased average/marginal estimates of exposure/treatment (A) on an outcome (Y) when A is administered at multiple time points and when a third factor (Z) is effected by A at time 0 and affects A at time 1
+- MSMs can help us get unbiased average/marginal estimates of exposure/treatment (A) on an outcome (Y) when A is administered at multiple time points and when a third factor (Z) is effected by A at time 0 and affects A at time 1.
 
@@
 2017, *An introduction to g methods*.
-It's a nicely written paper for those new to these causal inference methods and it also has SAS code for SAS users
+It's a nicely written paper for those new to these causal inference methods and it also has SAS code for SAS users.
 
-- In the example, we want to know the average effect of a binary treatment for HIV (A) on continous CD4 counts (Y).
+- In the example, we want to know the average effect of a binary treatment for HIV (A) on continuous CD4 counts (Y).
 Treatment is administered at multiple time points, with the only other covariate being a binary cutoff of HIV viral load (Z), where Z = 1 for viral load > 200 copies/ml.
@@
 ![](naimi_fig.png)
-
-## Testable research questions
-
-- There are a couple that come to mind:
-
-1. What is the average level of CD4 counts in those that received treatment at time 0 compared to those never treated?
 
@@
 
+## Testable research questions
+
+There are a couple that come to mind:
+
+1. What is the average level of CD4 counts in those that received treatment at time 0 compared to those never treated?
+
+2. What is the average level of CD4 counts in those that received treatment at time 1 compared to those never treated?
+
+3. What is the average level of CD4 counts in those treated at both time 0 and time 1 compared to those never treated?
+
 ## Standard estimates of the average effect of A on Y
 
-We'd like to know the average effect of treatment (at both time points)
+We'd like to know the average effect of treatment (at both time points).
 
-For reference, the 'true' average effect of A0 and A1 on Y is an increase of 50 CD4 count
+For reference, the 'true' average effect of A0 and A1 on Y is an increase of 50 CD4 count.
 
@@
 
-## G-formula accounting for time-dependent confounder
+## Reframing the question
 
-Now that we have our simulated data, we'll replicate the results Naimi et al.
-of marginal (overall) effect of HIV treatment on CD4 count
+- None of our various models using standard linear regression yielded the true value of treatment effect on CD4 count of 50.
 
-```{r, eval = F}
+- Let's ask the question what would the effect of treatment be on CD4 count if everyone were treated at both time points vs if no one was treated at both time points?
+
+- MSM's will let us estimate this question.
+
+## G-formula accounting for time-dependent confounder.
+
+Our question that we wish to answer with our MSM can be written as follows:
+
+$$
+\psi = E(Y^{a_0=1,a_1=1} - Y^{a_0=0,a_1=0})
+$$
+
+We can ask other "what if?" questions like "what if everyone was only treated at time 0 vs not treated at other time points?" etc., but we'll just ask this question for now.
+
+## Step 1: estimate standard relationships
+
+- Estimate exposures (A~0~,A~1~) relationship with the outcome (Y).
+- Estimate the effect of treatment (A~0~) on viral load (Z~1~).
+- Estimate the effect of viral load (Z~1~) on treatment (A~1~).
+
+```{r estimate associations}
 # g formula
@@
 a1_mod <- glm(A1 ~ Z1, data = td_sim_data, family = "binomial"(link="logit")) 
+```
 
-# natural course estimate ----
-nc_mu <- mean(predict(y_mod, newdata = td_sim_data, type = "response"))
+## Estimate Y where everyone is treated at both time points
 
+- We first estimate $$P(Z_1=z_1=1 | A_0 = a_0=1)$$
+
+- Followed by simulating a vector of new counter factual estimates of z~1~ values based on the prior probability of z~1~ = 1, given a~0~ = 1
+
+```{r z1 update}
 # fully exposed ----
@@
 z1_vec <- ifelse(rbernoulli(n = nrow(td_sim_data), z1_pr)==T,1,0)
+```
 
+## Alternate step
+
+- We can also estimate counter factual estimates of a~1~ if we wanted to ask "what if?" questions like "what if we only treated at time 1?"
+
+- But since we are asking treatment at both time points, this step is not needed.
+```{r a1 update}
 # update a1 
@@
 a1_vec <- ifelse(rbernoulli(n = nrow(td_sim_data), a1_pr)==T,1,0)
+```
+
+## Estimate Y under all exposed scenario.
+
+- Update our Z1 with z~1~ value in scenario where a~0~ = 1.
+- Updated our A1 and A0 value to all be 1.
+- If you had a new a~1~ vector, that could be added as well.
+
+```{r all exposed estimate}
 # update predicted Y where everyone is exposed at time 0 and 1, and accounting for Z1 | A0 =1 
@@
 all_exposed
+```
 
+## Repeat the same process for everyone being unexposed
+```{r unexposed estimate}
 # unexposed
@@
 all_unexposed
+```
+
+## Estimate the difference in scenarios
+
+We can now calculate our estimated average effect of treatment on Y where everyone is treated and no one is treated.
+
+```{r g comp difference}
 # diff
-all_exposed - all_unexposed
+g_diff <- all_exposed - all_unexposed
+g_diff
+```
 
+```{r g comp td formula, eval = F, echo = F}
+# hiding this for now, but general formula's will be needed if you want to bootstrap
 # creating a formula. will need dplyr
@@
 
-```{r iptw td, eval = F}
-MSModelR::iptw
+## IPTW with a time-dependent confounder
 
+First we estimate the propensity of A~0~, A~1~, and A~1~|Z~1~ using standard regression models
+
+```{r iptw td}
 # define propensity models (intercept only model without covariates)
@@
 pr_a1_z <- predict(pr_a1_z_mod, type = "response")
+```
 
+## Estimating weights
+
+```{r iptw weigts td}
 # create weights in dataframe
@@
 
-wt_table
+head(wt_table)
+```
 
+## Run a generalized model with the weights in the model
+
+We see we get the same result of an average effect of treatment on CD4 outcome is 50.
+
+```{r}
 # weighted model
@@
                 family = "gaussian")
-
-summary(iptw_mod)
-
-sum(iptw_mod$coefficients[names(iptw_mod$coefficients)=="A0" | 
+# calculate difference
+iptw_diff <- sum(iptw_mod$coefficients[names(iptw_mod$coefficients)=="A0" | 
   names(iptw_mod$coefficients)=="A1" | names(iptw_mod$coefficients)=="A0*A1"],1)
-
-```
-To make things easier, let's load some functions I've made for g-computation and IPTW.
-
-```{r, eval=F}
-# install custom package MSModelR
-# devtools::install_github("RyanGan/MSModelR")
-# load library
-library(MSModelR)
-
-test <- g.formula(data = td_sim_data, y = "Y", a = c("A0"), model.family = "gaussian",
-                  q.model = "A0+A1+A0*A1+Z1", estimate = "diff")
-
-iptw(data=td_sim_data, y = "Y", a = "A1", w = "Z1", estimate = "diff")
-
-?iptw
+# difference
+iptw_diff
 ```
 
-A benefit of these approaches over standard regression analyses is that they can provide estimates of the marginal effect of A on Y in the presence of interaction with A.
+## Comparison with our standard estimates
+
+We can see that both MSM terms give us our 'true' value, where our standard regression terms are off by a bit.
+
+Whether you think this is a big enough deal to use an MSM is up to you :)
+
+```{r std est w msm est, echo = F}
+std_est_msm_table <- std_est_table %>% 
+  add_row( model_params = c("G-Formula", "IPTW"), 
+           a_estimate = c(g_diff, iptw_diff))
+
+knitr::kable(std_est_msm_table, 
+  caption = "Standard and MSM estimates")
+```
+
+## MSMs when there is interaction with exposure
+
+Using an example out of **Implementation of G-Compuation on a Simulated Data Set: Demonstration of a Causal Inference Technique** by Snowden et al. 2011.
+
+This is example is probably a more practical reason to use MSMs to estimate a population-level effect.
+
+The example simulated an effect of a binary cutoff of high ozone (A) on FEV (Y), a continuous measure of pulmonary function.
+It also include sex (W1) as a confounding variable and an effect modifier for a binary race (W2) that has no base term.
+This means that race does not effect FEV, but that exposure is modified by race, attenuating the effect on FEV.
+
+The problem here is that we want to estimate a population-level effect for ozone on FEV, regardless of race.
+
+```{r simulate data, echo =F}
+# sample size
+n <- 300
+set.seed(285)
+simdata <- data.frame(W1 = rbinom(n, 1, 0.4), W2 = rbinom(n, 1, 0.5))
+# add simulated data of A
+simdata <- transform(simdata, A = rbinom(n, 1, (0.5+0.2*W1-0.3*W2)))
+# add simulated data of Y
+simdata <- transform(simdata, Y = rnorm(n, (3-0.5*A+W1+0.3*A*W2), 0.4))
+```
+
+## Standard estimation approach
+
+```{r snowden standard model}
+reg_mod <- glm(Y~A+W1+A:W2, data = simdata, family = gaussian)
+reg_mod
+```
+
+- If we want to know the causal effect of ozone (A) on FEV (Y), we have to some how account for the interaction with race (W2).
+
+- For our reference race, high level of ozone decrease FEV by 0.48.
+
+- For other race the effect of ozone decrease FEV by 0.18.
+
+- To estimate a population-level effect, perhaps we could average the two effects?
+
+- (0.48+0.18)/2 = 0.33
+
+## Let's try our MSM methods
+
+```{r msm with boot estimates, echo = F}
+# percentile method with broom
+# g formula
+set.seed(321)
+boot_msm <- simdata %>% bootstrap(1000) %>% 
+  do(tidy(MSModelR::g.formula(data = ., y = "Y", a = "A", q.model = "A+W1+A:W2",
+            model.family = "gaussian", estimate = "diff")))
+# iptw
+set.seed(123)
+boot_iptw <- simdata %>% bootstrap(1000) %>% 
+  do(tidy(MSModelR::iptw(data = ., y="Y",a="A",w="W1+W2",estimate="diff")))
+
+# create empty matrix
+estimates <- matrix(nrow = 2, ncol = 4, byrow = T)
+colnames(estimates) <- c("msm_estimate", "boot_median", "lower_95", "upper_95")
+
+# fill matrix for g formula 
+estimates[1,1] <- "G-Formula"
+estimates[1,2] <- round(quantile(boot_msm$x, 0.5),2)
+estimates[1,3] <- round(quantile(boot_msm$x, 0.025),2)
+estimates[1,4] <- round(quantile(boot_msm$x, 0.975),2)
+
+# fill matrix for iptw
+estimates[2,1] <- "IPTW"
+estimates[2,2] <- round(quantile(boot_iptw$x, 0.5),2)
+estimates[2,3] <- round(quantile(boot_iptw$x, 0.025),2)
+estimates[2,4] <- round(quantile(boot_iptw$x, 0.975),2)
+
+# convert matrix to dataframe
+est_df <- data.frame(estimates)
+
+knitr::kable(est_df, caption = paste0("MSM estimates with bootstrapped 95%CI"))
+```
+
+- G-formula estimate gives us the average value for each races of ozone on FEV.
+
+- IPTW is a little over this average estimate, but different from anyone race estimate alone.
+
+## MSMs
+
+Cons:
+
+- Producing estimates that may be interpreted as casual still requires assumptions like temporarily and no unmeasured confounding.
+
+- May be a lot of effort (and computationally expensive) to arrive at the same answer you'd get with a standard regression.
+
+- G-formula and IPTW don't handle continuous variables.
+
+Pros:
+
+- Most appropriate analytic method in some scenarios (time-dependent confounding).
+
+- Going through the whole process of figuring out causal structures (DAGs), checking assumptions, and implementing MSMs provides an excellent framework and tool for assessing causal relationships.
+
+## Cutting edge stuff
+
+- Doubly-robust estimation: combines both the q-model and propensity model and if even one of those models is specified correctly the estimate will be correct.
+
+- Targeted maximum likelihood (TMLE): uses an ensemble of machine learning algorithms to optimize the q-model and propensity model to control for confounders.
+
+- Structural nested models: old method that can handle continuous exposures but was difficult to implement; new packages for R are making this more accessible to average users.

### 2019-04-23 msm_introduction.Rmd
updated code to use rsample instead of broom for bootstrapped estimates

@@
 
-I will use the tidyverse and broom package for this presentation.
+I will use the tidyverse and rsample package for this presentation. Note, as of 2019, I use to use broom package, but bootstrapping functions have been depreciated in favor of rsample.
 
@@
 library(tidyverse)
-library(broom)
+library(rsample)
 ```
@@
 set.seed(321)
-boot_msm <- simdata %>% bootstrap(1000) %>% 
-  do(tidy(MSModelR::g.formula(data = ., y = "Y", a = "A", q.model = "A+W1+A:W2",
-            model.family = "gaussian", estimate = "diff")))
+
+# define simple substitution
+simple_sub <- function(split){
+  MSModelR::g.formula(data = analysis(split), y = "Y", a = "A", 
+                      q.model = "A+W1+A:W2", 
+                      model.family = "gaussian", estimate = "diff")
+  }
+# bootstrap using rsample package
+boot_msm <- simdata %>% 
+  bootstraps(times = 1000) %>% 
+  mutate(coef = map(splits, simple_sub)) %>% 
+  unnest(coef)
+
 # iptw
 set.seed(123)
-boot_iptw <- simdata %>% bootstrap(1000) %>% 
-  do(tidy(MSModelR::iptw(data = ., y="Y",a="A",w="W1+W2",estimate="diff")))
+
+iptw <- function(split){
+  MSModelR::iptw(data = analysis(split), 
+                 y="Y",a="A",w="W1+W2",estimate="diff")
+  }
+
+boot_iptw <- simdata %>% 
+  bootstraps(times = 1000) %>% 
+  mutate(coef = map(splits, iptw)) %>% 
+  unnest(coef)
 
@@
 estimates[1,1] <- "G-Formula"
-estimates[1,2] <- round(quantile(boot_msm$x, 0.5),2)
-estimates[1,3] <- round(quantile(boot_msm$x, 0.025),2)
-estimates[1,4] <- round(quantile(boot_msm$x, 0.975),2)
+estimates[1,2] <- round(quantile(boot_msm$coef, 0.5),2)
+estimates[1,3] <- round(quantile(boot_msm$coef, 0.025),2)
+estimates[1,4] <- round(quantile(boot_msm$coef, 0.975),2)
 
@@
 estimates[2,1] <- "IPTW"
-estimates[2,2] <- round(quantile(boot_iptw$x, 0.5),2)
-estimates[2,3] <- round(quantile(boot_iptw$x, 0.025),2)
-estimates[2,4] <- round(quantile(boot_iptw$x, 0.975),2)
+estimates[2,2] <- round(quantile(boot_iptw$coef, 0.5),2)
+estimates[2,3] <- round(quantile(boot_iptw$coef, 0.025),2)
+estimates[2,4] <- round(quantile(boot_iptw$coef, 0.975),2)
 

### 2019-07-16 g_comp_example_snowden.Rmd
update

@@
 
-Using code from Snowden paper to try out MSM models and maybe test out SuperLearner and Target Maximum Likelihood packages.
+Using code from Snowden paper to try out MSM models and SuperLearner and Target Maximum Likelihood packages.
 
@@
 library(tidyverse)
-library(broom)
+library(rsample)
 library(SuperLearner)
@@
 ```{r gcomp boot}
+# set funciton to handle splits for rsample package
+simple_sub <- function(split){
+  MSModelR::g.formula(data = analysis(split), y = "Y", a = "A", 
+                      q.model = "A+W1+A:W2", 
+                      model.family = "gaussian", estimate = "diff")
+}
+
 # percentile method with broom
 set.seed(321)
-boot_msm <- simdata %>% bootstrap(1000) %>% 
-  do(tidy(g_comp_func(data = ., y = "Y", a = "A", q_mod = "A+W1+A:W2",
-            mod_family = "gaussian", estimate = "diff")))
+boot_msm <- simdata %>% 
+  bootstraps(times = 1000) %>% 
+  mutate(coef = map(splits, simple_sub)) %>% 
+  unnest(coef)
 
@@
 estimates[,1] <- round(gcomp_est,3)
-estimates[,2] <- round(quantile(boot_msm$x, 0.5),3)
-estimates[,3] <- round(quantile(boot_msm$x, 0.025),3)
-estimates[,4] <- round(quantile(boot_msm$x, 0.975),3)
+estimates[,2] <- round(quantile(boot_msm$coef, 0.5),3)
+estimates[,3] <- round(quantile(boot_msm$coef, 0.025),3)
+estimates[,4] <- round(quantile(boot_msm$coef, 0.975),3)
 
@@
 ```{r iptw boot}
-# percentile method with broom
+# percentile method with rsample
 set.seed(321)
-boot_iptw <- simdata %>% bootstrap(1000) %>% 
-  do(tidy(iptw_func(data=., y="Y", a="A", w="W1+W2", estimate="diff")))
+
+iptw <- function(split){
+  iptw_func(data = analysis(split), 
+                 y="Y",a="A",w="W1+W2",estimate="diff")
+  }
+
+boot_iptw <- simdata %>% 
+  bootstraps(times = 1000) %>% 
+  mutate(coef = map(splits, iptw)) %>% 
+  unnest(coef)
 
@@
 estimates[,1] <- round(iptw_est,3)
-estimates[,2] <- round(quantile(boot_iptw$x, 0.5),3)
-estimates[,3] <- round(quantile(boot_iptw$x, 0.025),3)
-estimates[,4] <- round(quantile(boot_iptw$x, 0.975),3)
+estimates[,2] <- round(quantile(boot_iptw$coef, 0.5),3)
+estimates[,3] <- round(quantile(boot_iptw$coef, 0.025),3)
+estimates[,4] <- round(quantile(boot_iptw$coef, 0.975),3)
 
@@
 # register superlearner library
-sl_lib <- c("SL.glm", "SL.glm.interaction", "SL.gam", "SL.nnet", 
-            "SL.randomForest", "SL.step")
+sl_lib <- c("SL.glm", "SL.glm.interaction", "SL.gam", "SL.glmnet","SL.step")
 
@@
 head(cbind(simdata, expected_y_obsexp, expected_y_a1, expected_y_a0))
-
 ```
@@
   W = subset(simdata, select = -c(A,Y)),
-  Q.SL.library = c("SL.nnet", "SL.glm", "SL.randomForest", "SL.step",
-        "SL.glm.interaction"), 
-  g.SL.library = c("SL.nnet", "SL.glm", "SL.randomForest", "SL.step",
-        "SL.glm.interaction"),  V = 10)
+  Q.SL.library = c("SL.glm", "SL.step",
+        "SL.glm.interaction", "SL.xgboost"), 
+  g.SL.library = c("SL.glm", "SL.step",
+        "SL.glm.interaction", "SL.xgboost"),  V = 10)
 
-tmle_package_result
+summary(tmle_package_result)
 ```
 
-Wow.
 This was much faster and produced 95% CIs.

### 2021-08-10 causal_inference_examples.Rmd
update

---
title: "Causal Inference Examples"
author: "Ryan_Gan"
date: "January 19, 2017"
output: html_document
---

## Introduction

This markdown file was created to help me further understand key concepts in causal inference statistics to help in our causal inference papers. Recently (December 2016), Oleg Sofrygin created a pretty awesom package called 'simcausal'. Mark van der Laan (creator of the super learner package), and Romain Neugebauer are also authors on this package. This package makes creation of direceted acyclic graphs (DAGs) in a structural equations modeling (SEM) framework easy. Furthermore, it makes the simulation of data that adhere to your desired DAG very easy as well. I think this could eventually be useful for other people, but I'll need to be better about describing DAG theory.

*Disclaimer: Some of my DAG terminology may be off. I plan to come back and improve my discussions.*

```{r load libraries, warning=F, message=F}
library(simcausal) # simulation of causal structure data
library(broom) # for easy conversion of model results to dataframes
library(tidyverse) # data wrangling
```

## Simple Relationship

Let's start simple. Say you want to determine the effect of binary exposure "X" on binary outcome "Y". Let's create a DAG to show X affecting Y. In DAG terminology, the X and Y variables are sometimes called "nodes". The arrows connecting nodes are called "edges". 
The simcausal package even easily creates the DAG plot. I'm not familiar on all the plot options yet, so my DAGs may look pretty basic.

```{r x y dag, warning=F, message=F}
# using 'simcausal' package, first create an empty DAG using the appropriately
# named DAG.empty function.
d.1 <- DAG.empty()

# now using a SEM framework of notation, we define the relationship between 
# each node of the DAG
d.1 <- d.1 +
  # create node X
  # 'distr' defines the distribution we want to use. In this case, I want
  # a binary categorical variable with an equal distribution of 50% in the 
  # simulated data
  node("X", distr = "rcat.b1", prob = c(0.5, 0.5)) +  
  # create node Y
  # we use a Bernoulli distribution, which is equivalent to a binomial 
  # distribution when all X are independent, identically distribution (iid)
  # and random variables
  node("Y", distr = "rbern",
  # define the relationship between X and Y
  # lets say the baseline probability/risk for Y for unexposed X is ~.50
  # and the odds ratio (relative associatoin) for Y | X = exposed is 
  # roughly 2 times greater
  prob = plogis(0 + 0.69 * X))

# we now need to set the DAG we just created
d.1 <- set.DAG(d.1)

# now that this DAG is set, it's easy to plot
plotDAG(d.1, xjitter = 0, yjitter = 0,
        edge_attrs = list(width = 2, arrow.width = 2, arrow.size = 0.5),
        vertex_attrs = list(size = 24, label.cex = 0.8))

```

The DAG above shows a relationship between X and Y. In more specific "causal inference"" terminology we would say that "X affects Y".

We can also easily simulate data to analyze with the "simcausal" package.

```{r create data, warning=F, message=F}
# create simulated dataframe of DAG
# DAG option is the dag you want to simulate
# n is the sample size we want to simulate. Here we want 10,000 observatoins
# rndseed is the randomseed. Doesn't matter what number you define, but if you want
# to reproduce your data exactly, use the same seed
d.1_df <-  sim(DAG = d.1, n = 10000, rndseed = 321)

```

Of course we can always calculate our simple 2x2 sample by hand.

```{r hand cals, warning=F, message=F}
# find cell vals
xtabs(~X+Y, d.1_df)
# find odds ratio
odds_ratio <- (3997/1019)/(3333/1651) 
se <- sqrt((1/3997) + (1/1019) + (1/3333) + (1/1651))
lower_bound <- log(odds_ratio) - (se*1.96)
upper_bound <- log(odds_ratio) + (se*1.96)

names <- c("OR", "95_lower", "95_upper")
vals <- round(exp(c(log(odds_ratio), lower_bound, upper_bound)), 2)
or_df <- rbind(names, vals)
or_df

# find risk difference
risk_diff <- (3997/(3997+1019)) - (3333/(3333+1651))
risk_diff
# fill in standard error latter

```

Let's run a logistic regression on our simulated data to see if it has the properties we expect. I like to use the 'broom' (David Robinson) and 'tidyverse' (Hadley Wickham) to make data manipulation and output easier to read.

```{r xy mod, warning=F, message=F}

mod1 <- tidy(glm(Y~X, data = d.1_df, family="binomial"(link="logit")))
# model fits how we defined it
mod1

# calcualte 95%CI around estimate for X
estimate <- mod1[2,2]
lower_bound <- mod1[2,2] - (1.96*mod1[2,3])
upper_bound <- mod1[2,2] + (1.96*mod1[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
vals <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, vals)
or_df

```

Our simulated data fits our DAG. We may interpret the odds ratio in causal inference language as follows:
Exposure to X increased the odds of Y by ~2 times. This is consistent with our causal hypothesis of X affecting Y.

We could also estimate the risk difference in these data using a generalized linear regression model using binomial distribution with identity link.

```{r xy risk diff, warning=F, message=F}

risk_diff_mod <- tidy(glm(Y ~ X, data = d.1_df,family = "binomial"(link="identity")))
risk_diff_mod

```

Both our generalized linear regression models match what we would calculate by hand.

It's also important to note we could easily hypothesize that Y affects X. And those data may have the exact same relationship with each other.L et's draw a DAG and simulate.

```{r yx dag, warning=F, message=F}

dag2 <- DAG.empty()

# now using a SEM framework of notation, we define the relationship between 
# each node of the DAG
dag2 <- dag2 +
  node("Y", distr = "rcat.b1", prob = c(0.5, 0.5)) +  
  node("X", distr = "rbern",
  prob = plogis(0 + 0.69 * Y))

# we now need to set the DAG we just created
dag2 <- set.DAG(dag2)

# now that this DAG is set, it's easy to plot
plotDAG(dag2, xjitter = 0, yjitter = 0,
        edge_attrs = list(width = 2, arrow.width = 2, arrow.size = 0.5),
        vertex_attrs = list(size = 24, label.cex = 0.8))

df_2 <- sim(DAG = dag2, n = 10000, rndseed = 321)

# odds ratio 
or_mod <- tidy(glm(X ~ Y, data = df_2, family = "binomial"(link="logit")))
or_mod
# calcualte 95%CI around estimate for X
estimate <- or_mod[2,2]
lower_bound <- or_mod[2,2] - (1.96*or_mod[2,3])
upper_bound <- or_mod[2,2] + (1.96*or_mod[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
vals <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, vals)
or_df

# risk diff
risk_diff_mod <- tidy(glm(X ~ Y, data = df_2,family = "binomial"(link="identity")))
risk_diff_mod
```

Notice we get the exact same values in this newly simulated dataframe when we use the same parameters as before. The important lesson here is that statistical models make no distinction of direction of associaiton, and it's up to us as the modeler to build the best possible model we can that is consistent with our hypothesis.

As Pearl states, common words to express statistical relationships such as correlations or associations make no distinction of the direction of association. In many cases, we may want to use these terms. For exmaple, in the case of cross-sectional analyses. However, since we are talking about 'causal inference', it's important to note that we may say 'X affects/causes Y', but statistically, 'Y affects/causes X' could be just as likely given the data. This is where I think DAGs really help. They help visually represent our assumptions about our statistical model, given how our data were collected, our knowledge of the topic, etc.

## Confounding Bias

One area I believe the area of "causal inference" has really helped epidemiologic reasearch is by simplifying the biases that may be present in data to really just two major biases: confounding bias, and collider bias (selection bias). In a general sense confounding happens when a third variable/factor (which we'll call C for now) distorts the relationship between our X and Y association. I think the most common definition of confounding used in epidemiology courses today probably goes something like this:

*Confounding between X and Y occurs when there is an association between C and X, and there is an association between C and Y.* 
It may also include this extra piece: *And C is not on the causal pathway between X and Y.*

However, "causal inference" has a more specific definition of confounding:
*Confounding between X and Y occurs when C affects X and C affects Y.*

The DAG would look like this.

```{r confounding dag, warning=F, message=F}

con_dag <- DAG.empty()

# now using a SEM framework of notation, we define the relationship between 
# each node of the DAG
con_dag <- con_dag +
  node("C", distr = "rcat.b1", prob = c(0.5, 0.5)) +
  node("X", distr = "rbern", prob = plogis(0 + 1.5 * C)) +
  node("Y", distr = "rbern",
  prob = plogis(0 + 1.5 * C))

# we now need to set the DAG we just created
con_dag <- set.DAG(con_dag)

# now that this DAG is set, it's easy to plot
plotDAG(con_dag, xjitter = 0, yjitter = 0,
        edge_attrs = list(width = 2, arrow.width = 2, arrow.size = 0.5),
        vertex_attrs = list(size = 24, label.cex = 0.8))
```

Here is shows the X and Y are d-seperated, conditioned on C. We can see that in our model, where when we adjust for C, we would conclude there is no associatoin between X and Y.

```{r confound control, warning=F, message=F}

c_df <- sim(DAG = con_dag, n = 10000, rndseed = 321)

# unadjusted model
crude_mod <- tidy(glm(Y ~ X, data = c_df, family = "binomial"(link="logit")))

# calcualte 95%CI around estimate for X
estimate <- crude_mod[2,2]
lower_bound <- crude_mod[2,2] - (1.96*crude_mod[2,3])
upper_bound <- crude_mod[2,2] + (1.96*crude_mod[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
crude_vals <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, crude_vals)
or_df

# adjusted odds ratio 
adj_mod <- tidy(glm(Y ~ X + C, data = c_df, family = "binomial"(link="logit")))
adj_mod
# calcualte 95%CI around estimate for X
estimate <- adj_mod[2,2]
lower_bound <- adj_mod[2,2] - (1.96*adj_mod[2,3])
upper_bound <- adj_mod[2,2] + (1.96*adj_mod[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
adj_vals <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, adj_vals)
or_df

# risk diff
risk_diff_mod <- tidy(glm(Y ~ X + C, data = c_df,family = "binomial"(link="identity")))
risk_diff_mod

```

## Collider Bias

The second type of bias that occur within your data is collider bias. For epidemiologists, selection bias is a type of collider bias. A definition of collider bias would be:

*Collider bias occurs when X affects C, and where Y affects C.*

The using the same variables where X is our exposure, Y is our outcome, and C is some third variable that acts as a collider, the DAG would look like this.

```{r collider dag, warning=F, message=F}

col_dag <- DAG.empty()

# now using a SEM framework of notation, we define the relationship between 
# each node of the DAG
col_dag <- col_dag +
  node("X", distr = "rcat.b1", prob = c(0.5, 0.5)) +
  node("Y", distr = "rbern", prob = c(0.6, 0.4)) +
  node("C", distr = "rbern",
  prob = plogis(0 + 1.0 * Y + -0.5 *X))

# we now need to set the DAG we just created
col_dag <- set.DAG(col_dag)

# now that this DAG is set, it's easy to plot
plotDAG(col_dag, xjitter = 0, yjitter = 0,
        edge_attrs = list(width = 2, arrow.width = 2, arrow.size = 0.5),
        vertex_attrs = list(size = 24, label.cex = 0.8))
```

In this DAG, X and Y are d-seperated, unless we open up the path through collider C. Notice that the arrows leave from both X and Y *into* C. Thus C acts as a collider, opening up the path between X and Y if adjusted for.

```{r true model, warning=F, message=F}
# make collider dataframe
col_df <- sim(DAG = col_dag, n = 10000, rndseed = 321) %>% 
  # need to convert categorical Y to 1 or 0 to use with logistic
  mutate(Y = ifelse(Y == 1, 1, 0))

# true odds ratio 
true_mod <- tidy(glm(Y ~ X, data = col_df, family = "binomial"(link="logit")))
true_mod
# calcualte 95%CI around estimate for X
estimate <- true_mod[2,2]
lower_bound <- true_mod[2,2] - (1.96*true_mod[2,3])
upper_bound <- true_mod[2,2] + (1.96*true_mod[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
true_vals <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, true_vals)
or_df

```

In this case, our 'true' estimates (based on our DAG) show no association between X and Y. Which is also what our DAG shows, as X and Y are d-seperated.

However, what if we mistakenly think C is actually a confounder and decide to adjust for it in our model?

```{r collider model, warning=F, message=F}

# collider odds ratio 
col_mod <- tidy(glm(Y ~ X + C, data = col_df, family = "binomial"(link="logit")))
col_mod
# calcualte 95%CI around estimate for X
estimate <- col_mod[2,2]
lower_bound <- col_mod[2,2] - (1.96*col_mod[2,3])
upper_bound <- col_mod[2,2] + (1.96*col_mod[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
collider_vals <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, collider_vals)
or_df

```

If we adjust for C in this case, we have introduced collider bias. Our estimate is now biased as well. The lesson is that if the arrows go in to a variable, you may not want to adjust for such a variable. This could include mediating variables, which is a special case of collider bias. If the arrows leave the variable, then it should be adjusted for to control for confounding bias.

*Important reminder is that real data is messy, and these causal structures (DAGs) within data are not always apparent. Its good practice to draw multiple DAGs, test each node arch node in the DAG, and decide which DAG structure is more likely. This is the 'art' of model building, much like backwards or forwards selection, or some other model selection strategy. I prefer DAGs because it makes my assumptions about the questions I'm trying to ask explicit.*

Now what if we have a more complicated structure? Or we introduce collider bias on accident? Let's take a look at the classic M-bias example. I still don't know how or if you can control the locations of the nodes, so this DAG looks more like a star and not an M.

```{r m bias dag, warning = F, message = F}
m_dag <- DAG.empty()

# now using a SEM framework of notation, we define the relationship between 
# each node of the DAG
m_dag <- m_dag +
  node("C1", distr = "rbern", prob = c(0.5, 0.5)) +
  node("C2", distr = "rbern", prob = c(0.6, 0.4)) +
  node("M", distr = "rbern", prob = plogis(0 + 1.75 * C1 + 1.75 * C2)) +
  node("X", distr = "rbern",prob = plogis(0 + 2.5 * C1 + 3.5 * M)) +
  node("Y", distr = "rbern", prob = plogis(0 + 2.5 * C2 +3.5 * M))

# we now need to set the DAG we just created
m_dag <- set.DAG(m_dag)

# simulate m-dag data
m_df <- sim(DAG = m_dag, n = 100000, rndseed = 321)

# now that this DAG is set, it's easy to plot
plotDAG(m_dag, xjitter = 0, yjitter = 0,
        edge_attrs = list(width = 2, arrow.width = 2, arrow.size = 0.5),
        vertex_attrs = list(size = 24, label.cex = 0.8))

```

In the classic M bias paper (which I promise to cite), the question is asked, given the DAG, which variables need to be adjusted for in order to get the 'true' association between X and Y. The DAG shows M acting not only as a confounder on the association of X and Y, but also a collider in that while you may adjust for M, it would open up a colliding path between C1 and C2 that now both confound the association between X and Y. Only by adjusting for C1, C2, and M do we close all biasing paths.

We can see in our unadjusted model, that we might conclude falsely that there is an associatoin between X and Y.

```{r m bias model, warning = F, message = F}

m_mod <- tidy(glm(Y ~ X, m_df, family = "binomial"(link="logit")))
m_mod

# calcualte 95%CI around estimate for X
estimate <- m_mod[2,2]
lower_bound <- m_mod[2,2] - (1.96*m_mod[2,3])
upper_bound <- m_mod[2,2] + (1.96*m_mod[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
X <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, X)
or_df

```

If we adjust for just C1, we can see that this may be insufficient to account for all biasing paths as we would still conclude a significant association between X and Y.

```{r m bias adj C1 model, warning = F, message = F}

m_mod <- tidy(glm(Y ~ X + C1, m_df, family = "binomial"(link="logit")))
m_mod
# calcualte 95%CI around estimate for X
estimate <- m_mod[2,2]
lower_bound <- m_mod[2,2] - (1.96*m_mod[2,3])
upper_bound <- m_mod[2,2] + (1.96*m_mod[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
X <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, X)
or_df

```

Again, even though we adjust for C1 and C2 now, we would still conclude there is an associatoin between X and Y.

```{r m bias adj C1 C2 model, warning = F, message = F}

m_mod <- tidy(glm(Y ~ X + C1 + C2, m_df, family = "binomial"(link="logit")))
m_mod
# calcualte 95%CI around estimate for X
estimate <- m_mod[2,2]
lower_bound <- m_mod[2,2] - (1.96*m_mod[2,3])
upper_bound <- m_mod[2,2] + (1.96*m_mod[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
X <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, X)
or_df

```

Only by adjusting for C1, C2, and M do we close all biasing paths between X and Y, thereby creating d-seperation between X and Y.

```{r m bias adj C1 C2 M model, warning = F, message = F}

m_mod <- tidy(glm(Y ~ X + C1 + C2 + M, m_df, family = "binomial"(link="logit")))
m_mod
# calcualte 95%CI around estimate for X
estimate <- m_mod[2,2]
lower_bound <- m_mod[2,2] - (1.96*m_mod[2,3])
upper_bound <- m_mod[2,2] + (1.96*m_mod[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
X <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, X)
or_df

```

The M-bias DAG shows that there should be no associatoin between X and Y. However, in order for X and Y to be  d-seperated, we must adjust for C1, C2, and M. I'll add the citation for the M bias paper in here too. Also notice that I had to  make some really strong associations between the confounders (C1, C2), collider (M, e.g for associatoin between C1 and M, beta of 1.75 is really an OR of 5.75) and Y in order to make this example work. Anyone that's worked with real data knows that rarely will you see associations this strong.

The argument made by Greenland in his collider bias paper was that only by adjusting for C1, C2, and M, can the paths be closed, giving us the 'true' association. However, what if we just adjust for M only?

```{r m bias adj M model, warning = F, message = F}

m_mod <- tidy(glm(Y ~ X + M, m_df, family = "binomial"(link="logit")))
m_mod
# calcualte 95%CI around estimate for X
estimate <- m_mod[2,2]
lower_bound <- m_mod[2,2] - (1.96*m_mod[2,3])
upper_bound <- m_mod[2,2] + (1.96*m_mod[2,3])
# Odds ratio for the effect of X on Y
names <- c("OR", "95_lower", "95_upper")
X <- round(exp(c(estimate, lower_bound, upper_bound)), 2)
or_df <- rbind(names, X)
or_df

```

Heck, we get almost the exact same answer. Who has ever been asked this question and said you may just adjust for M and that would be good enough? I have :). I was also told I was 'wrong' because you need to adjust for C1 and C2 also. I argue based on modeling these simulate data that adjusting for M would be good enough. Even when the sample size is really large (here I made it one million) and if unrealistically strong associations were simulated, adjusting for M was still enough.

If anyone can create a scenario where we would miss the 'true' answer without fully adjusting for not only M, but C1 and C2 please let me know because adjusting for M always seemed to be enough. It's possible that perhaps if the variables were continous rather than binary, then it would be necissary to adjust for C1, C2, and not just M.

The lesson often told to us when we are taught DAGs is that only be identifying and accounting for all biasing paths in a complex DAG like this one, will we be able to arrive at a 'true' answer. The lesson I've learned from playing around with simmulated data of known causal structures is that  may suggest that often a simple DAG is sufficient to convey the most important biasing paths in your data.

## Things I'd like to do in the future

1. Show how matching in a case-control study increases efficiency, and how you can use the matchit package to quickly find controls.

2. Show how to perform mediation analyses.

3. Marginal sturctural models and inverse probability of treatment weight models. And finally see if these types of models can be used on continous/categorical data.

### 2021-08-10 causal_infernece_examples_v2.Rmd
update

---
title: "Causal inference methods examples part 2"
author: "Ryan Gan"
date: "8/7/2021"
output: html_document
---

```{r setup}
# loading r packages i'll use in this example
library(tmle) # targeted maximum likelihood
library(MASS)
library(xgboost) # gradient boosting with l2 regularization
library(glmnet) # lasso, ridge, and elastic regularization
library(gbm) # gradient boosting machiens
library(shapr) # explainable tree modles
library(tidyverse) # general data wrangling and plotting
library(mgcv) # for non-linear terms
```

Start by simulating a relationship between the treatment and a binary outcome. 
Adding two known confounders that affect both treatment and outcome and adding
some random normal variables with no association with either treatment or outcome
(although may happen by chance).

- Simple sim
- Interaction and non-linear sim
```{r sim}
# binary confounder example 
set.seed(100)
c1_x_association <- log(5)
c1_y_association <- log(5)
c2_x_association <- log(0.25)
c2_y_association <- log(0.25)

trt_out_association_true <- log(0.5) # true effect, protective effect

# sample size (note bias shrinks as sample size increases)
n <- 5000
# known confounders
c1 <- rbinom(n, 1, 0.6)
c2 <- rbinom(n, 1, 0.2)
age <- rnorm(n, mean = 55, sd = 5) # for non-linear function

# non-linear term on older patients less likely to get treatment

# 10 random noise variables to add to the model
rv_mat <- matrix(rnorm(n*10), ncol=10)
# give matrix some variable names 
colnames( rv_mat ) <- paste0('rv_', 1:10)

# relationship between confounders and treatment
trt <- rbinom(n, 1, (1 / (1+exp(-(c1*c1_x_association + c2*c2_x_association + 0.2*age + -0.0045*age^2)))))

# likelihood fun that includes relationship with treatment

lik_fun <- (
  1 / (1 + exp( -(trt*trt_out_association_true + c1*c1_x_association + c2*c2_x_association + 0.01*age ) ) )
  )

# binomial outcome based on the likelihood function
outcome <- rbinom(n, 1, lik_fun)

# put it all together in a dataframe to be used later
df <- data.frame( cbind(outcome, trt, c1, c2, age, rv_mat) )

```

2 x 2 and proportion table by treatment and outcome. In crude analysis, looks 
like both the relative risk and odds ratio suggest treatment increases likelihood
for the outcome. Note since our proportion of the outcome is high, the odds ratio
will not approximate the relative risk.

```{r 2x2}
tab <- xtabs(~ trt + outcome)
print(tab)

# crude relative risk
( tab[2,2] / sum(tab[2,]) ) / ( tab[1,2] / sum(tab[1, ]) )
# crude odds ratio
( tab[2,2] / tab[2,1])/ ( tab[1,2] / tab[1, 1] ) 
```

Estimation of crude associations with generalized linear models.
```{r crud_mods}
crude_trt_est <- broom::tidy( glm(outcome ~ trt, family = 'binomial') )[2, ]$estimate

# bias
print(paste0('Bias of True - Expected: ', round( crude_trt_est - trt_out_association_true , 2) ))
```

If we know the confounders of interest (based on previous studies or expert input),
a regression model works well enough for most cases.

```{r adj_mod}

# adjusted model formula with known confounders
adj_mod <- glm(outcome ~ trt + c1 + c2 + age, family = binomial(link='logit'), data = df)

# bias
print(
  paste0('Bias of True - Expected: ', 
         round( trt_out_association_true - adj_mod$coefficients['trt'], 2)
         )
  )
```

# G-Computation

G-computation (counterfactual estimate) first proposed by Robins in the epi 
literature sometime in the 1990s. In my opinion, this is Pearl's Do-calculus. 
Both state that if we specify a model `outcome | treatment, covariates` sufficiently
to account for confounding bias, then we have achieved condiitonal exchangability.
We can then estimate the marginal effect of treatment on the outcome by taking
the model and predicting as if everyone were treated, and predicting as if 
no one was treated. The average predicted score where everyone is treated can
then be compared to the average predicted score where no one was treated and 
that should be the marginal risk difference (in a binary outcome) or relative
ratio difference.

Using our adjusted model above, we can estimate this as follows:

```{r g_comp}
# copy dataframes
all_trt_df <- no_trt_df <- df
# set trt to all trt or no trt
all_trt_df$trt <- 1
no_trt_df$trt <- 0

# estimate using fit model and take the average of predictions
all_trt_pred <- predict(adj_mod, newdata = all_trt_df, type = 'response') 
no_trt_pred <- predict(adj_mod, newdata = no_trt_df, type = 'response')

# means for risk
trt_mean <- mean( all_trt_pred )
no_trt_mean <- mean( no_trt_pred )

# log odds ratio
mean_log_odds <- mean( log( ( all_trt_pred / (1 - all_trt_pred ) ) / (no_trt_pred / (1 - no_trt_pred)) ) )

# absolute reduction in risk
print(paste0( 'Risk difference: ', round( trt_mean - no_trt_mean , 2 ) ) )
# relative reduction in risk
print(paste0( 'Relative risk: ', round( trt_mean / no_trt_mean , 2 ) ) )
# odds ratio
gcomp_or <- ( trt_mean / ( 1 - trt_mean ) ) / (no_trt_mean / ( 1 - no_trt_mean ) )
print(paste0('Odds ratio: ', round( gcomp_or , 2)) )

# bias
print(
  paste0('Bias of True - Expected: ', 
         round( trt_out_association_true - mean_log_odds, 2)
         )
  )
```

All covs in model.

```{r g_comp_glm_incorrect}
inc_mod <- glm(outcome ~ trt + c1 + c2 + age + rv_1 + rv_2 + rv_3 + rv_4 + rv_5
               + rv_6 + rv_7 + rv_8 + rv_9 + rv_10, 
               family = binomial(link='logit'), data = df)

# estimate using fit model and take the average of predictions
all_trt_pred_inc <- predict(inc_mod, newdata = all_trt_df, type = 'response') 
no_trt_pred_inc <- predict(inc_mod, newdata = no_trt_df, type = 'response')

# means for risk
trt_mean_inc <- mean( all_trt_pred_inc )
no_trt_mean_inc <- mean( no_trt_pred_inc )

# log odds ratio
mean_log_odds_inc <- mean( 
  log( ( all_trt_pred_inc / (1 - all_trt_pred_inc ) ) / (no_trt_pred_inc / (1 - no_trt_pred_inc)) )
  )

# absolute reduction in risk
print(paste0( 'Risk difference: ', round( trt_mean_inc - no_trt_mean_inc , 2 ) ) )
# relative reduction in risk
print(paste0( 'Relative risk: ', round( trt_mean_inc / no_trt_mean_inc , 2 ) ) )
# odds ratio
gcomp_or_inc <- ( trt_mean_inc / ( 1 - trt_mean_inc ) ) / (no_trt_mean_inc / ( 1 - no_trt_mean_inc ) )
print(paste0('Odds ratio: ', round( gcomp_or_inc , 2)) )

# bias
print(
  paste0('Bias of True - Expected: ', 
         round( trt_out_association_true - mean_log_odds_inc, 2)
         )
  )

```

G-computation using lasso.

```{r g_comp_lasso}
# lasso
lasso_g_mod <- cv.glmnet(
  x = as.matrix( df[, 2:ncol(df)] ) , 
  y = df$outcome , 
  alpha = 1.0, # lasso penalty
  family = 'binomial'
)

# estimate using fit model and take the average of predictions
all_trt_pred_lasso <- predict(
  lasso_g_mod , 
  newx = as.matrix( all_trt_df[, 2:ncol(all_trt_df)] ), 
  s = "lambda.min", type = 'response'
)

no_trt_pred_lasso <- predict(
  lasso_g_mod , 
  newx = as.matrix( no_trt_df[, 2:ncol(no_trt_df)] ), 
  s = "lambda.min", type = 'response'
  )

# means for risk
trt_mean_lasso <- mean( all_trt_pred_lasso )
no_trt_mean_lasso <- mean( no_trt_pred_lasso )

# log odds ratio
mean_log_odds_lasso <- mean( 
  log( 
    ( all_trt_pred_lasso / (1 - all_trt_pred_lasso ) ) / (no_trt_pred_lasso / (1 - no_trt_pred_lasso))
    ) 
  )

# absolute reduction in risk
print(paste0( 'Risk difference: ', round( trt_mean_lasso - no_trt_mean_lasso , 2 ) ) )
# relative reduction in risk
print(paste0( 'Relative risk: ', round( trt_mean_lasso / no_trt_mean_lasso , 2 ) ) )
# odds ratio
gcomp_or_lasso <- ( ( trt_mean_lasso) / ( 1 - trt_mean_lasso ) ) / (no_trt_mean_lasso / ( 1 - no_trt_mean_lasso ) ) 
print(paste0('Odds ratio: ', round( gcomp_or_lasso , 2)) )

# bias
print(
  paste0('Bias of True - Expected: ', 
         round( trt_out_association_true - mean_log_odds_lasso, 2)
         )
  )

```

G-computation using xgboost.

```{r gcomp_xgb}
# xgb model
xgb_g_mod <- xgboost(
  data = as.matrix( df[, 2:ncol(df)] ),
  label = df$outcome, 
  max_depth = 4, eta = 0.01, nrounds = 5000, verbose = 0, nthread = 4, 
  objective = "binary:logistic"
)

# estimate using fit model and take the average of predictions
all_trt_pred_xgb <- predict(xgb_g_mod , newdata = as.matrix( all_trt_df[, 2:ncol(all_trt_df)] )) 
no_trt_pred_xgb <- predict(xgb_g_mod , newdata = as.matrix( no_trt_df[, 2:ncol(no_trt_df)] ))

# means for risk
trt_mean_xgb <- mean( all_trt_pred_xgb )
no_trt_mean_xgb <- mean( no_trt_pred_xgb )

# log odds ratio
mean_log_odds_xgb <- mean( 
  log( 
    ( all_trt_pred_xgb / (1 - all_trt_pred_xgb ) ) / (no_trt_pred_xgb / (1 - no_trt_pred_xgb))
    ) 
  )

# absolute reduction in risk
print(paste0( 'Risk difference: ', round( trt_mean_xgb - no_trt_mean_xgb , 2 ) ) )
# relative reduction in risk
print(paste0( 'Relative risk: ', round( trt_mean_xgb / no_trt_mean_xgb , 2 ) ) )
# odds ratio
gcomp_or_xgb <- ( ( trt_mean_xgb) / ( 1 - trt_mean_xgb ) ) / (no_trt_mean_xgb / ( 1 - no_trt_mean_xgb ) ) 
print(paste0('Odds ratio: ', round( gcomp_or_xgb , 2)) )

# bias
print(
  paste0('Bias of True - Expected: ', 
         round( trt_out_association_true - mean_log_odds_xgb, 2)
         )
  )

```

Plot showing bias from true
```{r gcomp_bias_plot}
est <- c('Unadjusted GLM', 'Correct Adj GLM', 'G-Comp Adj', 'G-Comp All Cov', 
         'G-Comp Lasso', 'G-Comp XGB')

g_coef <- c(crude_trt_est, adj_mod$coefficients['trt'], log( g_comp_or ), 
            log( gcomp_or_inc), log( gcomp_or_lasso ), log( gcomp_or_xgb ) )

g_df <- data.frame(
  estimation  = 'G-Computation', 
  method = est , 
  coef = g_coef
  ) %>% 
  mutate(
    method = factor( method , levels = method ), 
    odds_ratio = exp( coef )
  )

ggplot(g_df, aes( x = method, y = odds_ratio )) +
  geom_point() +
  geom_hline(yintercept = exp( trt_out_association_true ), color = 'red') +
  ylim( 0.25, 1.25 ) +
  theme_linedraw() +
  ylab('Odds Ratio') +
  xlab('Estimation Method')
```

## Propensity score

Another way to estimate is via propensity score. With this approach we model
the likelihood of being treated. Once we have a model, we can then get the 
predicted probability of being treated given the covariates and create weights.
Inverse predicted probability (IPTW) is common, but there are other weighting methods
that can account for some biases the arise with IPTW. Once we have a predicted
probability, we can create weights. I'm using a couple models below to illustrate
how this works using a naive linear model where the age effect is mis-specified,
a GAM model where a penalized spine is used for age, and an xgboost tree model
using all predictors (as if we knew nothing about the variables).

Note, we have a non-linear effect of age on treatment that should be accounted for.
If age is linear, the propensity might not fit as well. I'm not too worried about
overfitting because I want to try and predict as best I can on the observed
population and I'm not interested in predicting on an unseen population.

```{r propensity_models}
# age as a linear term
lin_pr_mod <- glm(
  as.formula(
    paste0('trt ~ ', paste(colnames(df)[3:ncol(df)], collapse = ' + ' ) )
    ), family = 'binomial', data = df
  )

# known gam model
gam_pr_mod <- gam(
    as.formula(
    paste0('trt ~ s(age) + ', paste(colnames(df)[c(3,4,6:ncol(df))], collapse = ' + ' ) )
    ), 
  family = 'binomial', data = df
  )

# lasso
lasso_pr_mod <- cv.glmnet(
  x = as.matrix( df[, 3:ncol(df)] ) , 
  y = df$trt , 
  alpha = 1.0, # lasso penalty
  family = 'binomial'
)

# xgb model
xgb_pr_mod <- xgboost(
  data = as.matrix( df[, 3:ncol(df)] ),
  label = df$trt, 
  max_depth = 4, eta = 0.01, nrounds = 1000, verbose = 0, nthread = 4, 
  objective = "binary:logistic"
)

# get predictions from models back
lin_pred <- predict(lin_pr_mod, newdata = df, type = 'response')

gam_pred <- predict(gam_pr_mod, newdata = df, type = 'response')

lasso_pred <- predict(lasso_pr_mod, newx = as.matrix( df[, 3:ncol(df)] ), s = "lambda.min", type = 'response')
colnames(lasso_pred) <- 'lasso_pred'

xgb_pred <- predict(xgb_pr_mod, newdata = as.matrix( df[, 3:ncol(df)] ))

```

Applying different weights. Average treatment effect (ATE), 
average treatment effect among treated (ATT), average treatment effect among
evenly match (ATM), average treatment effect of observed (ATO).

```{r create_ate_weights}
p_weights <- function(outcome, treatment, propensity_score) {
  # ate
  w_ate = (treatment / propensity_score) + ((1 - treatment) / (1 - propensity_score))

  # att
  w_att = ((propensity_score * treatment) / propensity_score) + 
      ((propensity_score * (1 - treatment)) / (1 - propensity_score))

  # atc
  w_atc = (((1 - propensity_score) * treatment) / propensity_score) + 
      (((1 - propensity_score) * (1 - treatment)) / (1 - propensity_score))

  # atm
  w_atm = pmin(propensity_score, 1 - propensity_score) / 
      (treatment * propensity_score + (1 - treatment) * (1 - propensity_score))

  # ato
  w_ato = (1 - propensity_score) * treatment + propensity_score * (1 - treatment)

  # dataframe with weights
  weight_df <- data.frame(
    outcome = outcome, 
    treatment = treatment,
    propensity = propensity_score, 
    w_ate = w_ate,
    w_att = w_att,
    w_atc = w_atc, 
    w_atm = w_atm, 
    w_ato = w_ato
    )

  return(weight_df)
}

# weight lists
lin_wt_df <- p_weights(outcome = df$outcome, treatment = df$trt, propensity_score = lin_pred)
gam_wt_df <- p_weights(outcome = df$outcome, treatment = df$trt, propensity_score = gam_pred)
lasso_wt_df <- p_weights(outcome = df$outcome, treatment = df$trt, propensity_score = as.vector( lasso_pred ) )
xgb_wt_df <- p_weights(outcome = df$outcome, treatment = df$trt, propensity_score = xgb_pred)

```

Plots of propensity scores.

```{r wt_plots}
d <- lin_wt_df %>% 
  tidyr::spread(treatment, propensity, sep = "_p")

ggplot(d) +
  geom_histogram(bins = 50, aes(treatment_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(treatment_p1, weight = w_ate, fill = "Treated"), alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = treatment_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = treatment_p0, weight = w_ate, y = -..count.., fill = "Not Treated"),
                 alpha = 0.5) + 
  scale_fill_manual('', values = c('blue', 'red') ) +
  ylab("Count") + 
  xlab("Propensity") +
  ggtitle('ATE weighting of propensity: linear model') +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs)  +
  theme_linedraw()

```

```{r lin_att}
ggplot(d) +
  geom_histogram(bins = 50, aes(treatment_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(treatment_p1, weight = w_att, fill = "Treated"), alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = treatment_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = treatment_p0, weight = w_att, y = -..count.., fill = "Not Treated"),
                 alpha = 0.5) + 
  scale_fill_manual('', values = c('blue', 'red') ) +
  ylab("Count") + 
  xlab("Propensity") +
  ggtitle('ATT weighting of propensity: linear model') +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs)  +
  theme_linedraw()
```
```{r lin_atm}
ggplot(d) +
  geom_histogram(bins = 50, aes(treatment_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(treatment_p1, weight = w_atm, fill = "Treated"), alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = treatment_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = treatment_p0, weight = w_atm, y = -..count.., fill = "Not Treated"),
                 alpha = 0.5) + 
  scale_fill_manual('', values = c('blue', 'red') ) +
  ylab("Count") + 
  xlab("Propensity") +
  ggtitle('ATM weighting of propensity: linear model') +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs)  +
  theme_linedraw()
```

Propensity models with weights.

```{r prop_wt_estimates}
# ate estimates
lin_prop_mod_ate <- glm(outcome ~ trt, data = lin_wt_df, family = 'quasibinomial', weights = w_ate) 
gam_prop_mod_ate <- glm(outcome ~ trt, data = gam_wt_df, family = 'quasibinomial', weights = w_ate) 
lasso_prop_mod_ate <- glm(outcome ~ trt, data = lasso_wt_df, family = 'quasibinomial', weights = w_ate) 
xgb_prop_mod_ate <- glm(outcome ~ trt, data = xgb_wt_df, family = 'quasibinomial', weights = w_ate)

# atm estimates
lin_prop_mod_atm <- glm(outcome ~ trt, data = lin_wt_df, family = 'quasibinomial', weights = w_atm) 
gam_prop_mod_atm <- glm(outcome ~ trt, data = gam_wt_df, family = 'quasibinomial', weights = w_atm) 
lasso_prop_mod_atm <- glm(outcome ~ trt, data = lasso_wt_df, family = 'quasibinomial', weights = w_atm) 
xgb_prop_mod_atm <- glm(outcome ~ trt, data = xgb_wt_df, family = 'quasibinomial', weights = w_atm) 
```

```{r propensity_estimates}
p_est <- c(
  'P-Score Lin ATE',
  'P-Score GAM ATE' , 
  'P-Score Lasso ATE',
  'P-Score XGB ATE' , 
  'P-Score Lin ATM',
  'P-Score GAM ATM' , 
  'P-Score Lasso ATM',
  'P-Score XGB ATM' 
  )

# propensity coef
p_coef <- c(
  # ate
  coef(lin_prop_mod_ate)['trt']
  , coef(gam_prop_mod_ate)['trt']
  , coef(lasso_prop_mod_ate)['trt']
  , coef(xgb_prop_mod_ate)['trt']
  # atm
  , coef(lin_prop_mod_atm)['trt']
  , coef(gam_prop_mod_atm)['trt']
  , coef(lasso_prop_mod_atm)['trt']
  , coef(xgb_prop_mod_atm)['trt'] 
  )

p_score_df <- data.frame(
  estimation = 'Propensity', 
  method = p_est, 
  coef = p_coef
  )  %>% 
  mutate(
    method = factor( method , levels = method ), 
    odds_ratio = exp( coef )
  )

```

## TMLE

This is a doubly-robust approach to estimating marginal effect that can use
machine learning models in an ensemble model.

```{r tmle}

# libraries to fit g and q formula in tmle
sl_libs <- c('SL.mean', 'SL.glm', 'SL.step', 'SL.glmnet', 'SL.gbm', 'SL.xgboost')

tmle_mod <- tmle(
  outcome, trt, as.matrix( df[, 3:ncol(df)] ) , family = 'binomial', 
  g.SL.library = sl_libs , 
  Q.SL.library = sl_libs , 
  V = 3, # use fewer cross validation folds, default it 10
)

```

```{r}
summary(tmle_mod)

```

```{r combind_metrics}
final_df <- g_df %>% 
  rbind(p_score_df) %>% 
  rbind(
    data.frame(
      estimation = 'TMLE',
      method = 'TMLE', 
      coef = -0.52012 , 
      odds_ratio = 0.59445
      )
    ) %>% 
  mutate(
    estimation = case_when(
      method %in% c('Unadjusted GLM', 'Correct Adj GLM') ~ 'Conditional Models', 
      TRUE ~ estimation
    )
  )

final_df
```

Final plot summarizing df.

```{r}
ggplot(
  data = final_df, 
  aes( x = factor(method, levels = method), y = odds_ratio , color = estimation) ) +
  geom_point() +
  coord_flip() +
  scale_x_discrete(
    limits = rev(factor(final_df$method, levels = final_df$method))
    )  +
  geom_hline(yintercept = exp( trt_out_association_true ), color = 'red') +
  ylim( 0.25, 1.25 ) +
  theme_linedraw() +
  ylab('Odds Ratio') +
  xlab('Estimation Method')
```

```{r}
  ggplot(data = median_impute_pois %>% filter(!str_detect(term, 'time_factor')), 
       aes( x = factor(term, levels = term), y = log(risk_ratio) ) ) +
  geom_point(position = position_dodge(width = 0.5), 
             color = 'darkblue') +
  geom_errorbar( aes( ymin = log(lower_95), ymax = log(upper_95) ), 
                 width = 0.3, 
                 position = position_dodge(width = 0.5),
                 color = 'darkblue') +
  geom_hline(yintercept = 0, color = 'red', linetype = 'dashed') +
  coord_flip() +
  scale_x_discrete(
    limits = rev(factor(results_table$term, levels = results_table$term))
    ) +
  xlab('Variable (continuous are standardized)') +
  ylab('Log(Risk Ratio)') +
  ggtitle('Poisson model') +
  theme_minimal()
```

### 2021-08-10 propensity_example.Rmd
update

---
title: "Propensity Example"
author: "Ryan Gan"
date: "7/6/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Propensity Example

Following Lucy D'Agostino McGowan's blog and dissertation work to understand newer weight methods for propensity scores.

https://livefreeordichotomize.com/2019/01/17/understanding-propensity-score-weighting/

Simulate data.

```{r}
set.seed(928)
n <- 1000
X <- mvtnorm::rmvnorm(n,
                      mean = c(0.5, 1),
                      sigma = matrix(c(2, 1, 1, 1), ncol = 2)
)

dat <- tibble(
  x_1 = X[, 1],
  x_2 = X[, 2],
  treatment = as.numeric(- 0.5 + 0.25 * x_1 + 0.75 * x_2 + rnorm(n, 0, 1) > 0),
  outcome = 2 * treatment + rnorm(n, 0, 1)
)
head(dat)
```

Fit propensity.

```{r}
dat <- dat %>%
  mutate(
    propensity_score = glm(treatment ~ x_1 + x_2, data = dat, family = "binomial") %>%
      predict(type = "response")
  )
```

Estimate weights from propensity.

```{r}
dat <- dat %>%
  mutate(
    w_ate = (treatment / propensity_score) + 
      ((1 - treatment) / (1 - propensity_score)),
    w_att = ((propensity_score * treatment) / propensity_score) + 
      ((propensity_score * (1 - treatment)) / (1 - propensity_score)),
    w_atc = (((1 - propensity_score) * treatment) / propensity_score) + 
      (((1 - propensity_score) * (1 - treatment)) / (1 - propensity_score)),
    w_atm = pmin(propensity_score, 1 - propensity_score) / 
      (treatment * propensity_score + (1 - treatment) * (1 - propensity_score)),
    w_ato = (1 - propensity_score) * treatment + 
      propensity_score * (1 - treatment)
  )
```

View propensity in mirrored display.

```{r}
d <- dat %>%
  tidyr::spread(treatment, propensity_score, sep = "_p")

ggplot(d) + 
  geom_histogram(bins = 50, aes(treatment_p1)) + 
  geom_histogram(bins = 50, aes(x = treatment_p0, y = -..count..)) + 
  ylab("count") + xlab("p") +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs) 
```

```{r}
ggplot(d) +
  geom_histogram(bins = 50, aes(treatment_p1), alpha = 0.5) + 
  geom_histogram(bins = 50, aes(treatment_p1, weight = w_ate), fill = "green", alpha = 0.5) + 
  geom_histogram(bins = 50, alpha = 0.5, aes(x = treatment_p0, y = -..count..)) + 
  geom_histogram(bins = 50, aes(x = treatment_p0, weight = w_ate, y = -..count..), fill = "blue", alpha = 0.5) + 
  ylab("count") + xlab("p") +
  geom_hline(yintercept = 0, lwd = 0.5) +
  scale_y_continuous(label = abs) 
```

### 2021-08-10 superlearner_vignette.Rmd
update

---
title: "SuperLearner Vignette"
author: "EC Polley"
date: "April 14, 2017"
output: html_document
---

## Purpose

Running through the SuperLearner Vignette to learn a bit about ensemble methods.

```{r libraries, echo = F}
library(SuperLearner)
library(glmnet)
library(kernlab)
library(KernelKnn)
```

Load the Boston housing data from MASS package. Seperating outcome from covariates. Creating training and confirmation datasets.

```{r boston data, echo = F}
# Setup example dataset.

# Load a dataset from the MASS package.
data(Boston, package = "MASS")

# Review info on the Boston dataset.
head(Boston)

# Check for any missing data - looks like we don't have any.
colSums(is.na(Boston))

# Extract our outcome variable from the dataframe.
outcome = Boston$medv

# Create a dataframe to contain our explanatory variables.
data = subset(Boston, select = -medv)

# Check structure of our dataframe.
str(data)

# If we had factor variables we would use model.matrix() to convert to numerics.

# Review our dimensions.
dim(data)

# Set a seed for reproducibility in this random sampling.
set.seed(1)

# Reduce to a dataset of 150 observations to speed up model fitting.
train_obs = sample(nrow(data), 150)

# X is our training sample.
X_train = data[train_obs, ]

# Create a holdout set for evaluating model performance.
# Note: cross-validation is even better than a single holdout sample.
X_holdout = data[-train_obs, ]

# Create a binary outcome variable: towns in which median home value is > 22,000.
outcome_bin = as.numeric(outcome > 22)

Y_train = outcome_bin[train_obs]
Y_holdout = outcome_bin[-train_obs]

# Review the outcome variable distribution.
table(Y_train, useNA = "ifany")
```

Load SuperLearner package and check available models. For maximum accuracy one might try at least the following models: glmnet, randomForest, XGBoost, SVM, and bartMachine. These should ideally be tested with multiple hyperparameter settings for each algorithm.

```{r sl library}
library(SuperLearner)
library(randomForest)
# review available models
listWrappers()

# look at glmnet model code
SL.glmnet

```

# Fit individual models

Let's fit 2 separate models: lasso (sparse, penalized OLS) and randomForest. We specify family = binomial() because we are predicting a binary outcome, aka classification. With a continuous outcome we would specify family = gaussian().

```{r rf and lasso models}
# Set the seed for reproducibility.
set.seed(1)

# Fit lasso model.
sl_lasso = SuperLearner(Y = Y_train, X = X_train, family = binomial(),
                        SL.library = "SL.glmnet")
sl_lasso

# Review the elements in the SuperLearner object.
names(sl_lasso)

# Here is the risk of the best model (discrete SuperLearner winner).
sl_lasso$cvRisk[which.min(sl_lasso$cvRisk)]

# Here is the raw glmnet result object:
str(sl_lasso$fitLibrary$SL.glmnet_All$object, max.level = 1)

# Fit random forest.
sl_rf = SuperLearner(Y = Y_train, X = X_train, family = binomial(),
                     SL.library = "SL.randomForest")
sl_rf
```

Risk is a measure of model accuracy or performance. We want our models to minimize the estimated risk, which means the model is making the fewest mistakes in its prediction. It's basically the mean-squared error in a regression model, but you can customize it if you want.

SuperLearner is using cross-validation to estimate the risk on future data. By default it uses 10 folds; use the cvControl argument to customize.

The coefficient column tells us the weight or importance of each individual learner in the overall ensemble. By default the weights are always greater than or equal to 0 and sum to 1. In this case we only have one algorithm so the coefficient has to be 1. If a coefficient is 0 it means that the algorithm isn't being used in the SuperLearner ensemble.

### Fit multiple models

Instead of fitting the models separately and looking at the performance (lowest risk), we can fit them simultaneously. SuperLearner will then tell us which one is best (discrete winner) and also create a weighted average of multiple models.

We include the mean of Y ("SL.mean") as a benchmark algorithm. It is a very simple prediction so the more complex algorithms should do better than the sample mean. We hope to see that it isn't the best single algorithm (discrete winner) and has a low weight in the weighted-average ensemble. If it is the best algorithm something has likely gone wrong.

```{r rf lasso and mean}
set.seed(1)
sl = SuperLearner(Y = Y_train, X = X_train, family = binomial(),
  SL.library = c("SL.mean", "SL.glmnet", "SL.randomForest"))
sl

# Review how long it took to run the SuperLearner:
sl$times$everything
```

Again, the coefficient is how much weight SuperLearner puts on that model in the weighted-average. So if coefficient = 0 it means that model is not used at all. Here we see that random forest is given the most weight, following by lasso.

### Predict on new data

Now that we have an ensemble let's predict back on our holdout dataset and review the results.
So we have an automatic ensemble of multiple learners based on the cross-validated performance of those learners, nice!

```{r predict on holdout}
# Predict back on the holdout dataset.
# onlySL is set to TRUE so we don't fit algorithms that had weight = 0, saving computation.
pred = predict(sl, X_holdout, onlySL = T)

# Check the structure of this prediction object.
str(pred)

# Review the columns of $library.predict.
summary(pred$library.predict)

# Histogram of our predicted values.
library(ggplot2)
qplot(pred$pred[, 1]) + theme_minimal()

# Scatterplot of original values (0, 1) and predicted values.
# Ideally we would use jitter or slight transparency to deal with overlap.
qplot(Y_holdout, pred$pred[, 1]) + theme_minimal()

# Review AUC - Area Under Curve
pred_rocr = ROCR::prediction(pred$pred, Y_holdout)
auc = ROCR::performance(pred_rocr, measure = "auc", x.measure = "cutoff")@y.values[[1]]
auc

```

AUC can range from 0.5 (no better than chance) to 1.0 (perfect). So at 0.96 we are looking pretty good!

### Customize a model hyperparameter

Hyperparameters are the configuration settings for an algorithm. OLS has no hyperparameters but essentially every other algorithm does.

There are two ways to customize a hyperparameter: make a new learner function, or use create.Learner().

Let's make a variant of random forest that fits more trees, which may increase our accuracy and can't hurt it (outside of small random variation).

```{r sl mod of rf}
# Review the function argument defaults at the top.
SL.randomForest

# Create a new function that changes just the ntree argument.
# (We could do this in a single line.)
# "..." means "all other arguments that were sent to the function"
SL.rf.better = function(...) {
  SL.randomForest(..., ntree = 3000)
}

set.seed(1)

# Fit the CV.SuperLearner.
# We use V = 3 to save computation time; for a real analysis use V = 10 or 20.
cv_sl = CV.SuperLearner(Y = Y_train, X = X_train, family = binomial(), V = 3,
                        SL.library = c("SL.mean", "SL.glmnet", "SL.rf.better", "SL.randomForest"))

# Review results.
summary(cv_sl)
```

Looks like our new RF is not improving performance. This implies that the original 500 trees had already reached the performance plateau - a maximum accuracy that RF can achieve unless other settings are changed (e.g. max nodes).

For comparison we can do the same hyperparameter customization with create.Learner().

```{r custom learner}
# Customize the defaults for randomForest.
learners = create.Learner("SL.randomForest", params = list(ntree = 3000))

# Look at the object.
learners

# List the functions that were created
learners$names

# Review the code that was automatically generated for the function.
# Notice that it's exactly the same as the function we made manually.
SL.randomForest_1

set.seed(1)

# Fit the CV.SuperLearner.
# We use V = 3 to save computation time; for a real analysis use V = 10 or 20.
cv_sl = CV.SuperLearner(Y = Y_train, X = X_train, family = binomial(), V = 3,
                        SL.library = c("SL.mean", "SL.glmnet", learners$names, "SL.randomForest"))

# Review results.
summary(cv_sl)
```

We get exactly the same results between the two methods of creating a custom learner.

### Test algorithm with multiple hyperparameter settings

The performance of an algorithm varies based on its hyperparamters, which again are its configuration settings. Some algorithms may not vary much, and others might have far better or worse performance for certain settings. Often we focus our attention on 1 or 2 hyperparameters for a given algorithm because they are the most important ones.

For randomForest there are two particularly important hyperparameters: mtry and maximum leaf nodes. Mtry is how many features are randomly chosen within each decision tree node - in other words, each time the tree considers making a split. Maximum leaf nodes controls how complex each tree can get.

Let's try 3 different mtry options.

```{r hyperparameter settings}
# sqrt(p) is the default value of mtry for classification.
floor(sqrt(ncol(X_train)))

# Let's try 3 multiplies of this default: 0.5, 1, and 2.
mtry_seq = floor(sqrt(ncol(X_train)) * c(0.5, 1, 2))
mtry_seq

learners = create.Learner("SL.randomForest", tune = list(mtry = mtry_seq))

# Review the resulting object
learners

# Check code for the learners that were created.
SL.randomForest_1
SL.randomForest_2
SL.randomForest_3

set.seed(1)

# Fit the CV.SuperLearner.
# We use V = 3 to save computation time; for a real analysis use V = 10 or 20.
cv_sl = CV.SuperLearner(Y = Y_train, X = X_train, family = binomial(), V = 3,
                        SL.library = c("SL.mean", "SL.glmnet", learners$names, "SL.randomForest"))

# Review results.
summary(cv_sl)
```

We see here that mtry = 3 performed a little bit better than mtry = 1 or mtry = 7, although the difference is not significant. If we used more data and more cross-validation folds we might see more drastic differences. A higher mtry does better when a small percentage of variables are predictive of the outcome, because it gives each tree a better chance of finding a useful variable.

Note that SL.randomForest and SL.randomForest_2 have the same settings, and their performance is very similar - statistically a tie. It's not exactly equivalent due to random variation in the two forests.

A key difference with SuperLearner over caret or other frameworks is that we are not trying to choose the single best hyperparameter or model. Instead, we usually want the best weighted average. So we are including all of the different settings in our SuperLearner, and we may choose a weighted average that includes the same model multiple times but with different settings. That can give us better performance than choosing only the single best settings for a given algorithm, which has some random noise in any case.

#### Stopping here for now; the rest is about parallel computing

### 2021-10-29 causal_infernece_examples_v2.Rmd
update

@@
 author: "Ryan Gan"
-date: "8/7/2021"
-output: html_document
+date: "2021-10-19"
+output: 
+  html_document:
+    code_folding: hide
+    toc: true
+    toc_depth: 2
 ---
@@
 ```{r setup}
-# loading r packages i'll use in this example
+# loading r packages 
 library(tmle) # targeted maximum likelihood
@@
 Adding two known confounders that affect both treatment and outcome and adding
-some random normal variables with no association with either treatment or outcome
-(although may happen by chance).
+some random normal variables with no association with either treatment or outcome (although may happen by chance).
 
-- Simple sim
-- Interaction and non-linear sim
 ```{r sim}
@@
 
-# G-Computation
+# G-computation
 
-G-computation (counterfactual estimate) first proposed by Robins in the epi 
-literature sometime in the 1990s. In my opinion, this is Pearl's Do-calculus. 
-Both state that if we specify a model `outcome | treatment, covariates` sufficiently
-to account for confounding bias, then we have achieved condiitonal exchangability.
-We can then estimate the marginal effect of treatment on the outcome by taking
-the model and predicting as if everyone were treated, and predicting as if 
-no one was treated. The average predicted score where everyone is treated can
-then be compared to the average predicted score where no one was treated and 
-that should be the marginal risk difference (in a binary outcome) or relative
-ratio difference.
+G-computation (a.k.a simple substitution, counterfactual estimate) first proposed by Robins in the epi literature sometime in the 1990s.
+In my opinion, this is Pearl's Do-calculus.
+Both state that if we specify a model `outcome | treatment, covariates` sufficiently to account for confounding bias, then we have achieved conditional exchangeability.
+We can then estimate the marginal effect of treatment on the outcome by taking the model and predicting as if everyone were treated, and predicting as if no one was treated.
+The average predicted score where everyone is treated can then be compared to the average predicted score where no one was treated and that should be the marginal risk difference (in a binary outcome) or relative 6ratio difference.
 
@@
 
-All covs in model.
+All covs in model; saturated model.
 
-```{r g_comp_glm_incorrect}
-inc_mod <- glm(outcome ~ trt + c1 + c2 + age + rv_1 + rv_2 + rv_3 + rv_4 + rv_5
-               + rv_6 + rv_7 + rv_8 + rv_9 + rv_10, 
+```{r g_comp_glm_satorrect}
+saturated_mod <- glm(outcome ~ trt + c1 + c2 + age + rv_1 + rv_2 + rv_3 + rv_4
+                     + rv_5 + rv_6 + rv_7 + rv_8 + rv_9 + rv_10, 
                      family = binomial(link='logit'), data = df)
@@
 # estimate using fit model and take the average of predictions
-all_trt_pred_inc <- predict(inc_mod, newdata = all_trt_df, type = 'response') 
-no_trt_pred_inc <- predict(inc_mod, newdata = no_trt_df, type = 'response')
+all_trt_pred_sat <- predict(saturated_mod, newdata = all_trt_df, type = 'response') 
+no_trt_pred_sat <- predict(saturated_mod, newdata = no_trt_df, type = 'response')
 
 # means for risk
-trt_mean_inc <- mean( all_trt_pred_inc )
-no_trt_mean_inc <- mean( no_trt_pred_inc )
+trt_mean_sat <- mean( all_trt_pred_sat )
+no_trt_mean_sat <- mean( no_trt_pred_sat )
 
 # log odds ratio
-mean_log_odds_inc <- mean( 
-  log( ( all_trt_pred_inc / (1 - all_trt_pred_inc ) ) / (no_trt_pred_inc / (1 - no_trt_pred_inc)) )
+mean_log_odds_sat <- mean( 
+  log( ( all_trt_pred_sat / (1 - all_trt_pred_sat ) ) / (no_trt_pred_sat / (1 - no_trt_pred_sat)) )
   )
@@
 # absolute reduction in risk
-print(paste0( 'Risk difference: ', round( trt_mean_inc - no_trt_mean_inc , 2 ) ) )
+print(paste0( 'Risk difference: ', round( trt_mean_sat - no_trt_mean_sat , 2 ) ) )
 # relative reduction in risk
-print(paste0( 'Relative risk: ', round( trt_mean_inc / no_trt_mean_inc , 2 ) ) )
+print(paste0( 'Relative risk: ', round( trt_mean_sat / no_trt_mean_sat , 2 ) ) )
 # odds ratio
-gcomp_or_inc <- ( trt_mean_inc / ( 1 - trt_mean_inc ) ) / (no_trt_mean_inc / ( 1 - no_trt_mean_inc ) )
-print(paste0('Odds ratio: ', round( gcomp_or_inc , 2)) )
+gcomp_or_sat <- ( trt_mean_sat / ( 1 - trt_mean_sat ) ) / (no_trt_mean_sat / ( 1 - no_trt_mean_sat ) )
+print(paste0('Odds ratio: ', round( gcomp_or_sat , 2)) )
 
@@
   paste0('Bias of True - Expected: ', 
-         round( trt_out_association_true - mean_log_odds_inc, 2)
+         round( trt_out_association_true - mean_log_odds_sat, 2)
          )
@@
 
-Plot showing bias from true
+Plot showing bias from true.
+
 ```{r gcomp_bias_plot}
@@
 
-g_coef <- c(crude_trt_est, adj_mod$coefficients['trt'], log( g_comp_or ), 
-            log( gcomp_or_inc), log( gcomp_or_lasso ), log( gcomp_or_xgb ) )
+g_coef <- c(crude_trt_est, adj_mod$coefficients['trt'], log( gcomp_or ), 
+            log( gcomp_or_sat), log( gcomp_or_lasso ), log( gcomp_or_xgb ) )
 
@@
 
-Plots of propensity scores.
+Plots of propensity scores weighting schemes.
 
@@
 ```
+
 ```{r lin_atm}
@@
     method = factor( method , levels = method ), 
-    odds_ratio = exp( coef )
+    odds_ratio = exp( coef ), 
+    percent_bias = coef - log(0.5)
   )
@@
 
-## TMLE
+Plot for propensity score estimates difference from the true estimate.
+
+```{r pscore_plot}
+ggplot(p_score_df, aes( x = method, y = odds_ratio )) +
+  geom_point() +
+  geom_hline(yintercept = exp( trt_out_association_true ), color = 'red') +
+  ylim( 0.25, 1.25 ) +
+  theme_linedraw() +
+  ylab('Odds Ratio') +
+  xlab('Estimation Method') +
+  theme(axis.text.x = element_text(angle = 60, vjust = 0.5))
+```
+
+## Targeted-maximum likelihood estimation (TMLE)
 
@@
 
-```{r}
+```{r tmle_summary}
+
 summary(tmle_mod)
@@
 
-Final plot summarizing df.
+Final plot summarizing bias of estimates compared to true value.
 
-```{r}
+```{r summary_plot}
 ggplot(
@@
 ```
-
-```{r}
-  ggplot(data = median_impute_pois %>% filter(!str_detect(term, 'time_factor')), 
-       aes( x = factor(term, levels = term), y = log(risk_ratio) ) ) +
-  geom_point(position = position_dodge(width = 0.5), 
-             color = 'darkblue') +
-  geom_errorbar( aes( ymin = log(lower_95), ymax = log(upper_95) ), 
-                 width = 0.3, 
-                 position = position_dodge(width = 0.5),
-                 color = 'darkblue') +
-  geom_hline(yintercept = 0, color = 'red', linetype = 'dashed') +
-  coord_flip() +
-  scale_x_discrete(
-    limits = rev(factor(results_table$term, levels = results_table$term))
-    ) +
-  xlab('Variable (continuous are standardized)') +
-  ylab('Log(Risk Ratio)') +
-  ggtitle('Poisson model') +
-  theme_minimal()
-```
